---
title: "Week 11 Exercises"
author: "Brad McNeney"
date: '2017-11-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use the `Heart` data set, split into roughly equal-size
training and test sets. There is a short quiz on Canvas named
`Week 11 Tutorial Quiz` for you to fill out after doing these questions.

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Heart.csv")
Heart <- read.csv(uu,row.names=1)
Heart <- na.omit(Heart)
# Some variables are factors. Some are not but probably should be:
unlist(lapply(Heart,is.factor))
unlist(lapply(Heart,{function(x) length(unique(x))}))
library(dplyr)
Heart <- mutate(Heart,
                RestECG = factor(RestECG),
                Slope = factor(Slope),
                Ca = factor(Ca))
n.train <- ceiling(nrow(Heart)/2)
set.seed(1)
train <- sample(1:nrow(Heart),size=n.train,replace=FALSE)
```

1. Fit a GAM to the training data with `AHD` as the binary 
response and all other variables as explanatory variables.
Fit smooths, `s()`, for the variables
`Age`, `RestBP`, `Chol`, `MaxHR` and `Oldpeak`. Use this 
model to predict `AHD` in the test set using the 
probability threshold 0.5. Calculate the misclassification 
rate. Enter your rate on the Canvas quiz.


2. Fit a boosted tree model to the training data with `AHD` as the binary 
response and all other variables as explanatory variables. Use
500 trees and the default shrinkage value. Before calling 
`gmb()`, set the random seed to 1.
Use the boosted tree 
model to predict `AHD` in the test set using the 
probability threshold 0.5. 
(Note: The `predict()` method for `gbm` objects
requires that you specify the number of trees.)
Calculate the misclassification 
rate. Enter your rate on the Canvas quiz.


3. Fit a support vector machine with radial kernel
to the training data with 
`AHD` as the binary 
response and all other variables as explanatory variables. 
Use the `gamma=1/2` and `cost=1`.
Use the SVM to predict `AHD` in the test set.
Calculate the misclassification 
rate. Enter your rate on the Canvas quiz.

