---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 2: Statistical Learning, R supplement'
author: "Brad McNeney"
date: '2017-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## What is R?

- R is an open-source environment for statistical computing and graphics.
- Started in the mid-1990's at Auckland University
- Now maintained by a team of experts called the R Development Core Team
- A "packages" system allows any user to bundle R code, data and examples together.
    + Load packages with `library()`
- R and R packages are distributed through the Comprehensive R Archive Network (CRAN). 
- SFU has a CRAN mirror at <http://cran.stat.sfu.ca>

## What does "environment" mean?
- R is a fully-functioning programming environment with all the usual constructs, such as 
    - conditionals (if-then-else), 
    - loops 
    - user-defined functions.
- In addition there are built-in facilities for 
    - data input, storage, manipulation, and output
    - optimization, matrix computation, etc., 
    - random number generation,
    - data analysis and graphics.
- "Base" R is good, but it is the package system that makes R great.


## Starting R
- Start R by starting RStudio. 
- The "Console" window is where you can type your commands.
- However, it is good practice to open an R script, type your
commands in the script, and then submit the commands to the R console.
    - `Session -> Set Working Directory` to set the working directory
    - `File -> New File -> R Script` to open a new R script
    - type your commands into the script
    - put your cursor on the line you want to submit and hit `Ctrl-enter`
- Save your script for later use.
- More on the RStudio interface at <https://support.rstudio.com/hc/en-us/sections/200107586-Using-RStudio>

## R Cheatsheets

* Use Google to find one that works for you.
* This one looks OK to me:
<http://github.com/rstudio/cheatsheets/raw/master/source/pdfs/base-r.pdf>

## R objects
- In R, data structures and functions are all referred to 
as "objects".
- Objects are created with the assignment operator `<-`; e.g.,
`x <- 1`.
    - The objects a user creates from the R console are contained in
    the user's workspace, called the global environment. 
    - Use `ls()` to see a list of all objects in the workspace. 
    - Use `rm(x)` to remove object `x` from the workspace.


## R Data Structures

- Focus on four common data structures: atomic vectors, lists, matrices and data frames.
- Atomic vectors and lists are 1d, while matrices and data frames are 2d objects
- R has no true scalars; e.g., in `x<-1`, `x` is a vector of length one.
- R also has an array data structure for higher dimensional elements that we will not discuss.
- Use `str()` to see the structure of an object


## Vectors

- Vectors can be either atomic or list
    - atomic vectors must be comprised entirely of logical, integer, double (numeric) or character elements
    - lists can be comprised of multiple data types
- Data vectors can be created with `c()` or `list()`:

\footnotesize

```{r}
avec <- c(50,200,77)
lvec <- list(50,200,77,c("grey","thin"))
```

## Combining vectors


- Use `c()` to combine vectors

\footnotesize

```{r}
c(avec,c(100,101))
c(lvec,TRUE)
```


## Factors

- The statistical concept of a factor is important in experimental 
design. 
- Factors are implemented in R as atomic vectors with attributes
`class` and `levels`:

\footnotesize

```{r}
trt <- factor(c("drug1","placebo","placebo","drug2"))
attributes(trt)
str(trt)
```
- The levels are coded numerically (1, 2 and 3)
with assigned labels ordered alphabetically
("drug1", "drug2" and "placebo")


## Subsetting vectors and extracting elements

- Subset with `[` or by name:
```{r}
lvec[c(1,3)] # same as lvec[c("age","height")]
```
- Extract individual elements with `[[`, or `$` for named objects:
```{r}
lvec[[4]]
lvec$hair
```


## Subsetting and assignment

- You can combine subsetting and assignment to change the value of 
vectors

\footnotesize

```{r}
avec
avec[2] <- 210
avec
```

## Matrices and data frames
- Though both 2d objects, matrices and data frames
are different enough that we will need to discuss them 
separately.
- The elements of a matrix must all be of the same type.
- Data frames are essentially lists where each list element
has the same length. Thus data frames can include columns
of varying type.

## Matrices


- Matrices can be created with the `matrix()` function as in 

\footnotesize

```{r}
A <- matrix(1:4,nrow=2,ncol=2) 
A
```

\normalsize

- Here `1:4` is the same as `c(1,2,3,4)`
- The default is to read the data vector into the matrix column-by-column. To 
read row-by-row instead use the `byrow=TRUE` argument:

\footnotesize

```{r}
A <- matrix(1:4,nrow=2,ncol=2,byrow=TRUE) 
A
```

## Combining matrices

- Combine matrices with `rbind()` and `cbind()`:

\footnotesize

```{r}
rbind(A,matrix(c(5,6),nrow=1,ncol=2))
cbind(A,A)
```


## Subsetting matrices 

- Subset with `[` and a comma to separate rows from columns:

\scriptsize

```{r}
A[1,1]
A[1,]
A[,1]
```

\small

- When a subsetting operation leads to a vector, the dimension 
of the object is "dropped" from 2 to 1. To prevent this use
`drop=FALSE`:

\scriptsize

```{r}
A[1,,drop=FALSE]
```


## Extracting elements from matrices 

```{r}
A[1,1]
```


##  Data frames
\small

- Data frames (class `data.frame`) are the usual way to 
store data in R.
    - Rows are intended to be observational units, columns variables
    - Implemented as a list (columns are list elements), but also behave like a matrix in terms of combining and subsetting.
- Create with `data.frame`:

\footnotesize


```{r}
set.seed(1)
n <- 4
x <- 1:n; y <- rnorm(n,mean=x,sd=1) # multiple commands separated by ;
dd <- data.frame(x=x,y=y) # like making a list
str(dd)
```

## Subsetting and combining data frames

\small

- Can subset columns like a list:

\scriptsize

```{r}
dd$x 
```

\small

- Can subset columns/rows and combine like matrices; e.g.,

\scriptsize


```{r}
dd[1:2,]
zz = data.frame(z=runif(4))
cbind(dd,zz)
```


## Logical operators
\small

- `!` is NOT
- `&` and `&&` are AND, with `&` acting vector-wise and `&&` acting 
on scalars
- `|` and `||` are OR, with `|` acting vector-wise and `||` acting 
on scalars
- Make sure you understand the following:

\footnotesize

```{r}
x <- c(TRUE,TRUE,FALSE); y <- c(FALSE,TRUE,TRUE)
!x ; x&y ; x&&y ; x|y ; x||y
```

## Relational operators
\small

- Relational operators can be used to compare values in atomic vectors
    - See `help("Comparison")`
- `>` is greater than, `>=` is greater than or equal
- `<` is less than, `<=` is less than or equal
- `==` is equal and `!=` is not equal
- Make sure you understand the following:

\footnotesize

```{r}
x <- 1:3; y <- 3:1
x>y ; x>=y ; x<y ; x<=y ; x==y ; x!=y
```


## Subsetting vectors with logical expressions
\small

- Can subset with logicals and `[`:

\footnotesize


```{r}
avec
avec>100
avec[avec>100]
avec[avec>50 & avec<100]
```

## Subsetting matrices with logical expressions

- Can also subset matrices, but results may not be as expected:

\footnotesize

```{r}
A
A>1
A[A>1] # coerces to a vector
```

## Missing values

- R has a special data code for missing data: `NA`
- Test for and set missing values with `is.na()`

\footnotesize


```{r}
avec
is.na(avec)
is.na(avec) <- 2 
avec
```


## R functions: Example 

\small

```{r}
f <- function(x) {
  return(x^2)
}
f
```



## Reading Data: Native format


- Use `save()` to save R objects to an "R Data" file.
    - `save.image()` is short-hand to save all objects 
    in the workspace

\footnotesize

```{r}
x <- rnorm(100); y <- list(a=1,x=x)
save(x,y,file="test.RData") # Or .rda, or ...
```

\normalsize

- Load R Data files into the workspace with `load()`.

\footnotesize

```{r}
load("test.RData")
file.remove("test.RData") 
```

## Reading Table Format Files

- `read.table()` is the main function for reading 
tabular data from plain-text files.
    - `read.csv()` and `read.delim()` are basically 
    `read.table()` with defaults for reading comma- and tab-
    delimited files.
- `write.table()`, `write.csv()` and `write.delim()` are
the analogous functions for writing tabular data

\footnotesize

```{r}
write.table(matrix(1:9,3,3),file="test.txt")
test <- read.table("test.txt")
file.remove("test.txt")
test
```

## Reading files from a URL

- `load()`, `read.table()`, etc. can read data from a URL.

\scriptsize

```{r}
baseURL <- "http://people.stat.sfu.ca/~mcneney/Teaching/Stat452/"
rdURL <- url(paste0(baseURL,"Data/PorschePrice.rda"))
load(rdURL)
head(PorschePrice)
csvURL <- url(paste0(baseURL,"Data/PorschePrice.csv"))
PorschePrice <- read.csv(csvURL)
```

## `stringsAsFactors`

- Reading columns that include characters
in as factors is controlled by a global option in your R session
called `stringsAsFactors`, set to `TRUE` by default.
- If you want to set to `FALSE` for an R session type
`options(stringsAsFactors = FALSE)` into the Console.
- An alternative is to over-ride the default in the
call to `read.table()`:

\footnotesize

```{r}
exURL <- url(paste0(baseURL,"Data/Ex1_1_4.txt"))
ex2 <- read.table(exURL,header=TRUE,sep="\t",
                  stringsAsFactors=FALSE) 
```

## Viewing Data: `print()`, `View()` and `edit()`

- `print()` prints R objects
    - This function is "generic", meaning that 
    it will try to find the 
    specific function to print specific objects
    (e.g., `print.data.frame`).
- `View()` launches a new window (or RStudio tab)
to view a data frame and `edit()` launches a 
data editor.

## Graphics

* "Base" graphics in R is good, but `ggplot()` is better.
* We could spend a lot of time on `ggplot()`, but will 
just learn what we need as we go.
- ggplot2 cheatsheet at [https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf]
- Wickham (2009) ggplot2: Elegant graphics for data analysis, Chapters 4 and 5.
- Chang (2012) R graphics cookbook. Available at  [http://www.cookbook-r.com/Graphs/]

