---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 6, Part 3: Dimension Reduction Methods'
author: "Brad McNeney"
date: '2018-10-14'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3.5,fig.width=5)
```

## Reduced Dimension Regression

* Transform predictors $X_1,\ldots,X_p$ to
a lower-dimension set $Z_1,\ldots,Z_M$, for
$M<p$.
    + The $Z_m$'s are taken to be linear combinations of the 
    $X_j$'s: 
    $$Z_m = \sum_{j=1}^p \phi_{jm} X_j$$
* Fit a linear model to $Z_1,\ldots,Z_M$
$$Y = \theta_0 + \sum_{m=1}^M \theta_m Z_m + \epsilon.$$
Compare to the linear model
$$Y = \beta_0 + \sum_{j=1}^p \beta_m X_m + \epsilon.$$
    + Fewer regression coefficients ($M+1 < p+1$).

## Lower Dimension, Constraint on $\beta$'s

* As shown on pages 229,230 of the text,
the lower-dimension model implies coefficients
in the original model of the form
$$\beta_j = \sum_{m=1}^M \theta_m \phi_{jm}$$
* Thus the $p$ $\beta$s are constrained to be
functions of $M$ underlying $\theta$s.
    + Different form of constraints from those in
    ridge regression and the lasso (recall the 
    second view of these as constrained maximization).
* Introduction of a constraint is 
another way to view the bias/variance trade-off:
    + constraints mean lower variance, but higher bias
    on parameter estimates, which translates into 
    lower variance/higher bias for predictions.
    
    
## Methods for Dimension Reduction

* Principal components -- low-rank approximation of the 
$X$ data matrix
* Partial least squares -- explain $X$ by latent variables

## Principal Components Analysis (PCA)

* More details on PCA to follow in Chapter 10.
* First centre each variable by subtracting its mean.
* Then, think of principal components (PCs) as
new coordinates for the data vectors.
    + The first PC is the direction of greatest variation,
    + The second PC is the direction of second-greatest
    variation, orthogonal to the first,
    + And so on.

## PCs for Advertising Data


* Text Figure 6.14:
The green line is the first PC, the blue 
line the second.

\begin{center}
\includegraphics[height=2.5in]{Fig6_14.pdf}
\end{center}

## PCs as Linear Combinations of $X$'s

* We won't go into the details of how the 
linear combinations are derived.
* In the advertising example, the first PC is 
$$Z_1 = 0.838 X_1 + 0.544 X_2$$
where $X_1$ is population centred by its mean
and $X_2$ is advertising expenditure centred by its mean.
* The coefficients of the linear combination,
$\phi_{11}= 0.838$ and $\phi_{12}=0.544$,
are called the first principal component *loadings*.

## Principal Component Scores

* Projecting each point onto the PCs gives the 
PC scores.
    + Projecting a data vector 
    onto a line means finding the point
    on the line closest to the vector.

* Text Figure 6.15: Black x's are the first PC score
for each observation, distance of each purple 
dot from the green line is the second PC score.

\vspace*{-.5in}

\begin{center}
\includegraphics[height=2in]{Fig6_15.pdf}
\end{center}

## High-Dimensional Example: Genes Reflect Geography

* First 2 PCs from 
197,146 genetic markers on 1,387 European individuals
(Novembre _et al._ 2008)

\begin{center}
\includegraphics[height=2.5in]{EuropeanGenStr.pdf}
\end{center}


## PCs and PC Scores for the Credit Data

\scriptsize

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Credit.csv")
Credit <- read.csv(uu,row.names=1)
head(Credit,n=3)
X <- model.matrix(Balance ~ ., data=Credit)
X <- X[,-1] # Remove intercept
X <- scale(X) # Centre and scale
pcs <- prcomp(X)
```

##

\scriptsize

```{r}
plot(pcs)
```

## Loadings for First Two PCs

\scriptsize

```{r}
pcs$rotation[,1:2]
```

##

\scriptsize

```{r}
plot(pcs$x[,1],pcs$x[,2],xlab="First PC",ylab="Second PC")
```


## Principal Components Regression (PCR)

* Take $Z_1,\ldots,Z_M$ to be the first $M$ 
PC scores.
    + $M$ can be chosen by cross-validation
    to minimize estimated test set error.
* The idea is that a handful of PCs might 
explain the variation in $X$ **and** the 
relationship between $X$ and $Y$.

## PCR on the Credit Data

\scriptsize

```{r}
library(pls) # install.packages("pls")
set.seed(123)
cfit <- pcr(Balance ~ ., data=Credit,scale=TRUE,
            validation="CV")
```

## Summary

\tiny

```{r}
summary(cfit)
```

\normalsize

* Note: `RMSEP` is root mean squared error of prediction,
the square root of the MSE.

## Plot the Root MSE of Prediction

\scriptsize

```{r}
validationplot(cfit)
```

## Extract $\hat{\beta}$'s

* These are the estimates of the coefficients of the $X$'s,
$$\beta_j = \sum_{m=1}^M \theta_m \phi_{jm}$$

\scriptsize

```{r}
coef(cfit,ncomp=10)
```

## Partial Least Squares (PLS) versus PCR

* Statistical learning methods that use
the response are said to be "supervised", while
those that do not are "unsupervised".
* PCR does unsupervised selection of the 
transformed features $Z_1,\ldots,Z_M$.
* By contrast, PLS is supervised (sketch of details below).
* No clear winner between PCR and PLS. 
    + Supervised dimension reduction may reduce
    bias by identifying features that are truly 
    related to $Y$.
    + However, supervising "... has the potential
    to increase variance," (text, page 238)

## PLS Directions

* The loadings for the first PLS direction, $Z_1$ are the 
coefficients from the simple linear 
regression of $Y$ on each $X_j$.
* The loadings for the second PLS direction are 
coefficients from the simple linear regression of
the *adjusted* variable
$Y-\hat{Y}$ on the adjusted $X_j-\hat{X}_j$, where
$\hat{Y}$ and $\hat{X}_j$ are from regressions on $Z_1$.
    + The residuals are the information in the variables
    not explained by $Z_1$.
* The loadings for the third PLS direction are 
coefficients from the simple linear regression of
the adjusted variable
$Y-\hat{Y}$ on the adjusted $X_j-\hat{X}_j$, where
$\hat{Y}$ and $\hat{X}_j$ are from regressions on $Z_1$
**and** $Z_2$.
    + The residuals are the information in the variables
    not explained by $Z_1$ and $Z_2$.
* And so on.


## PLS on the Credit Data

\scriptsize

```{r}
cfit <- plsr(Balance ~ ., data=Credit,scale=TRUE,
            validation="CV")
```


## Summary

\scriptsize

```{r}
summary(cfit)
```

## Plot the Root MSE of Prediction

\scriptsize

```{r}
validationplot(cfit)
```

## Extract $\hat{\beta}$'s

\scriptsize

```{r}
coef(cfit,ncomp=4)
```