---
title: "Homework 3"
author: "name1 and name2 and name3"
date: '2017-10-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Question 1 (Chapter 6, #8, parts (a)-(e), 10 marks)


(a) (1 mark)
(Note: You should set your random seed, for reproducibility.)

(b) (1 mark)


(c) (3 marks)
For the "best model obtained", you should 
use one that is parsimonious and close to
the consensus best according tht the three
selection criteria.

You don't **have** to create a data frame. 
`regsubsets()` can take a design matrix and
response vector, just like `lm.fit()` and 
`glmnet()`. If you do decide to create a data frame,
the following hint may be of use:
```{r}
library(leaps)
pmax <- 10
Xmat <- matrix(NA,nrow=n,ncol=pmax)
for(i in 1:pmax) {
  Xmat[,i] <- X^i
}
colnames(Xmat) <- paste0("X.",1:pmax)
dat <- data.frame(Y,Xmat)
```


(d) (2 marks) 


(e) (3 marks)

```{r}
library(glmnet)
lambdas <- 10^{seq(from=-2,to=5,length=100)}
cv.lafit <- cv.glmnet(Xmat,Y,alpha=1,lambda=lambdas) 
plot(cv.lafit)
la.best.lam <- cv.lafit$lambda.1se
la.best.lam
la.best <- glmnet(Xmat,Y,alpha=1,lambda=la.best.lam)
coef(la.best)
```


## Question 2 (Ch6, #9, 12 marks)

(a) (0 marks)
To make everyone's results comparable, please
select your test set with the following.

```{r}
library(ISLR)
data(College)
dim(College) # 777 rows, use 111 as test
set.seed(1)
testset <- sample(1:777,size=111)
College.test <- College[testset,]
College.train <- College[-testset,]
```

(b) (2 marks)

(c) (2 marks)

(d) (2 marks)

(e) (2 marks)

(f) (2 marks)

(g) (2 marks)


## Question 3 (Ch7, #6, 8 marks)

(a) (5 marks)

(b) (3 marks)
