---
title: "Homework 1 Solutions"
author: "Brad McNeney"
date: '2017-09-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1 (Chapter 2, #1, 8 marks)

(a) (2 marks) Flexible is better (1 mark), because it 
can use the abundance of data points (large $n$) to give a good fit.

(b) (2 marks) Flexible is worse (1 mark) because it
takes more data ($n$) to fit.

(c) (2 marks) Flexible is better (1 mark) because it 
will have the flexibility to fit the non-linear relationship.

(d) (2 marks) Flexible is worse (1 mark) because it
will try to fit the error.


## Question 2 (Chapter 2, #2, 6 marks)

(a) (2 marks) This is a regression problem and we 
are most interested in inference (1 mark). 
$n=500$, $p=3$.

(b) (2 marks) This is a classification problem and we
are  most interested in prediction (1 mark).
$n=20$, $p=13$.

(c) (2 marks) This is a regression problem and we are
most interested in prediction (1 mark).
$n=52$, $p=3$.

## Question 3 (Chapter 2, #9, 8 marks)

```{r}
library(ISLR) 
data(Auto)
head(Auto)
```

(a) (1 mark) From the help page `help("Auto")`, we see
that origin and name are qualitative and the other seven
are quantitative. 
Note: A variable that takes on only a few values is **discrete**,
but that doesn't mean it must be qualitative. (Though
all qualitative variables are discrete.)
`cylinders` is discrete, but is quantitative.
The test for a quantitative variable is whether it makes sense
to do arithmetic. 

(b) (1 mark) 
Note: My hint regarding `summarize()` didn't work! Fixed
on Sept. 18.

```{r}
myr <- function(x) {
  out <- range(x)
  names(out) <- c("min","max")
  out
}
apply(Auto[,1:7],MARGIN=2,FUN=myr)
```


(c) (1 mark)

```{r}
meansd <- function(x) {
  out<-c(mean(x),sd(x))
  names(out) <- c("mean","SD")
  out
}
apply(Auto[,1:7],MARGIN=2,FUN=meansd)
```


(d) (2 marks; one for subsetting, one for the summaries)
Note: My original code to subset the `Auto` dataframe 
didn't work. Fixed on Sept. 18.

```{r}
AutoSubset <- Auto[c(1:9,86:nrow(Auto)),1:7]
rangeMeanSd <- function(x) {
  c(myr(x),meansd(x))
}
apply(AutoSubset,MARGIN=2,FUN=rangeMeanSd)
```

(e) (2 marks) One mark for some pairwise 
scatterplots of the quantitative variables
and comment on the relationships. You might
start by using the `pairs()` function for plots and 
`cor()` for correlations among the quantitative
variables. One mark for a sensible comment.
For example:
The variables `cylinders`, `displacement`, `horsepower`
and `weight` are positively correlated with each other
and negatively correlated with `mpg`.

You could also have done individual
scatterplots, and perhaps used colors to indicate the `origin`.
Note that in `ggplot()` qualitative variables should 
be coerced to factors.

```{r}
pairs(Auto[,1:7])
cor(Auto[,1:7])
library(dplyr)
Auto <- mutate(Auto,fOrigin = factor(origin,labels=c("American","European","Japanese")))
               
library(ggplot2)
ggplot(Auto,aes(x=displacement,y=mpg,color=fOrigin)) +
  geom_point() + geom_smooth(method="lm",se=FALSE)
```

(f) (1 marks) You should have included some 
plots or numerical summaries to show that `mpg` looks
associated with all of the quantitative variables
(and in some cases looks quite strongly associated).
