---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 1: Introduction'
author: "Brad Mcneney"
date: '2017-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## What is Statistical Learning?

* Tools for learning from data.
* Multiple regression is an example of such a tool:
    + We propose a linear model $Y = \beta_0 + X_1 \beta_1 + \ldots + X_p \beta_p + \epsilon$, and fit the model.
    + We may interpret fitted coefficients, or use them 
    to obtain predictions.
    + Such a problem is said to be "supervised" because of 
    the response variable $Y$, viewed as an "output"
    that is influenced by the "inputs" $X_1,\ldots,X_p$.
    
## Example: The `Wage` data

\scriptsize

```{r}
library(ISLR)
data(Wage)
head(Wage)
```

##

\scriptsize

```{r,fig.height=3,fig.width=4}
library(ggplot2)
ggplot(Wage,aes(x=age,y=wage)) +
  geom_point() + geom_smooth()
```


##

\scriptsize

```{r,fig.height=3,fig.width=4}
ggplot(Wage,aes(x=year,y=wage)) + geom_point() + geom_smooth(method="lm")
```

##

\scriptsize

```{r,fig.height=3,fig.width=4}
library(ggplot2)
ggplot(Wage,aes(x=education,y=wage)) +
  geom_boxplot()
```

##

\scriptsize

```{r}
wfit <- lm(wage ~ age + I(age^2) + year + education, data=Wage)
summary(wfit)$coefficients
```


## Other tools

* We will study non-linear methods for supervised
learning, and methods appropriate
to "unsupervised" problems, where there is no response 
variable.

## Notation

* There are $n$ distinct observations.
* The random response for the $i$th individual 
is $Y_i$ and observed value is $y_i$; $i=1\ldots,n$.
* There are $p$ explanatory variables $X=(X_1, \ldots, X_p)$.
* The measured value of the
$j$th explanatory variable on the $i$th observation
is denoted $x_{ij}$


## More Notation and Simple Matrix Algebra

* Will to discuss in tutorial for those who need a refresher
or quick intro.

## R and RStudio

* You will need to install both R and the RStudio interface,
**and** you will need to create an RStudio "project"
based on the class GitHub repository.
    + See the computer software "getting started" page on
    canvas.
    + Also, Will to discuss in tutorial.
* Also be sure to 
install the `tidyverse` and
`ISLR` packages using the RStudio `Tools` menu, or
from the command line

```{r,eval=FALSE}
install.packages(c("tidyverse","ISLR")
```

