---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 8, Part 2: Bagging and Random Forests'
author: "Brad McNeney"
date: '2018-10-31'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Example: The Hitters Data Set

* Example tree grown, but not pruned.

\scriptsize

```{r}
library(ISLR)
data(Hitters)
library(dplyr)
Hitters <- mutate(Hitters,lSalary = log(Salary)) %>% 
  select(-Salary) %>%
  na.omit()
library(tree)
t1 <- tree(lSalary ~ .,data=Hitters)
```

##

\scriptsize

```{r,fig.height=5,fig.width=5}
plot(t1)
text(t1,cex=.5)
```

## Bootstrapping 

* A bootstrap sample from $n$ observations is obtained 
by drawing $n$ observations, with replacement, from 
the sample

\scriptsize

```{r}
set.seed(543)
n <- 6
x <- round(rnorm(n),1)
sort(x)
sort(sample(x,size=n,replace=TRUE)) # bootstrap sample 1, 
sort(sample(x,size=n,replace=TRUE)) # bootstrap sample 2
sort(sample(x,size=n,replace=TRUE)) # bootstrap sample 3
```

## Empirical Distribution

* The data-generating process in our example is a normal distribution.
* The empirical distribution of our sample is the distribution
under which each sampled value is equally likely.
* Resampling amounts to sampling from the empirical distribution.
* If empirical close to data-generating distribution, 
resampled samples are like new samples from the data-generating 
distribution.

## Example: Tree from a Bootstrap Sample

\scriptsize

```{r}
# Use sample_n() from dplyr
n <- nrow(Hitters)
bHitters <- sample_n(Hitters,size=n,replace=TRUE)
t1b <- tree(lSalary ~ .,data=bHitters)
```

##

\scriptsize

```{r,fig.height=5,fig.width=5}
plot(t1b)
text(t1b,cex=.5)
```


## Bootstrap Aggregation (Bagging)

* For a general statistical learning method that 
provides an estimate, $\hat{f}(x)$ of the model $f(x)$,
we can aggregate over estimates fit to bootstrap samples.
* If $\hat{f}^b(x)$ is the fitted model from the $b$th 
bootstrap sample, the aggregate estimator is
$$\hat{f}_{\rm avg}(x) = \frac{1}{B}\sum_{b=1}^B \hat{f}^b(x),$$
where $B$ is the number of bootstrap samples.
* Bagging has been found to be useful for improving 
predictions from decision trees.
    + Grow each tree "deep" and don't prune.
    + Average predictions over trees.
  

## Bias-Variance Tradeoff

* The deeply-grown trees have high variance, but low bias.
* Averaging many deep trees is an attempt to reduce the
variance.
* However, if trees are highly correlated (e.g., similar first 
few splits) then averaging does not reduce variance.
    + Return to this point when we discuss random forests.
    


## Bagging for Classification

* When the outcome $Y$ is categorical, one approach to
the aggregate prediction of category is the "majority vote":
    + Record the predicted class from each tree (each 
    bootstrap sample) and take the most common prediction
    as the aggregate prediction.
    
## Estimating Test Error with Out-of-Bag (OOB) Observations

* On average about 2/3 are and 1/3 of observations are not resampled
    + See small bootstrap samples of size 6 above.
* Those not resampled are called out-of-bag (OOB)
* Regression trees: The OOB prediction for the $i$th observation
is the average of its predicted values from trees grown
on boostrap samples that did not include $i$.
    + The OOB MSE is the average squared error between  observed
    values and their OOB predictions.
* Classification trees: The OOB prediction for the $i$th observation
is the majority vote from trees grown on bootstrap samples that did
not include $i$.
    + The OOB classification error is the proportion of misclassified
    OOB predictions. 
* OOB errors are valid because the predictions use only trees not
fit to the corresponding observation.
    
## Variable Importance

* There isn't one tree to interpret now.
* Need an overall measure of the importance of a variable.
* For a single tree, a variable's importance can be measured
by the amount that the node homogeneity (RSS for regression,
Gini for classification) is decreased by splitting on the variable. 
    + Homogeneity of tree including variable 
    minus homogeneity if variable was not available
    to split on.
    

## Example: Heart Data

\scriptsize

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Heart.csv")
Heart <- read.csv(uu,row.names=1)
Heart <- na.omit(Heart)
dim(Heart) # Train on 2/3, test on 1/3
set.seed(1)
train <- sample(1:nrow(Heart),size=2*nrow(Heart)/3,replace=FALSE)
t1 <- tree(AHD ~ ., data=Heart,subset=train)
```

## 

\scriptsize

```{r,fig.height=5,fig.width=6}
plot(t1)
text(t1,cex=.5,pretty=0)
```

##

\scriptsize

```{r}
cv.t1 <- cv.tree(t1,FUN=prune.misclass) # use class'n err for CV
plot(cv.t1) # Suggest size 7
```

## 

\scriptsize

```{r}
t1.best <- prune.tree(t1,best=7)
summary(t1.best)
hpred <- predict(t1.best,newdata=Heart[-train,],type="class")
tt <- table(hpred,Heart[-train,]$AHD)
tt
sum(tt[row(tt) != col(tt)])/sum(tt)
```

## Bagging on Heart Data

\scriptsize

```{r}
library(randomForest)
set.seed(1)
bag.heart <- randomForest(AHD ~ ., data=Heart,subset=train,
                          mtry=13,importance=TRUE) #m=p is bagging
bag.heart
```

##

\scriptsize

```{r,fig.height=4.3,fig.width=7}
varImpPlot(bag.heart) # or importance(bag.heart)
```

\normalsize

* See `help("importance")` for a definition of accuracy.

## Classification Error for Bagged Trees

\scriptsize

```{r}
bag.hpred <- predict(bag.heart,newdata=Heart[-train,],type="class")
tt <- table(bag.hpred,Heart[-train,]$AHD)
tt
sum(tt[row(tt) != col(tt)])/sum(tt)
```

## Random Forests

* Bootstrap datasets may yield correlated trees.
    + E.G., if we have one or two strong predictors, all trees
    might share the first few splits.
* Random forests avoid this by restricing the predictors 
to a random subset.
    + New subset at each split.
    + Subset size? Text: "... typically we choose  $m\approx\sqrt{p}$".

## Random Forests on Heart Data

\scriptsize

```{r}
rf.heart <- randomForest(AHD ~ ., data=Heart,subset=train,
                          mtry=sqrt(13),importance=TRUE) #m<p for RF
rf.heart
```

##

\scriptsize

```{r}
rf.hpred <- predict(rf.heart,newdata=Heart[-train,],type="class")
tt <- table(rf.hpred,Heart[-train,]$AHD)
tt
sum(tt[row(tt) != col(tt)])/sum(tt)
```

##

\scriptsize

```{r,fig.height=4.5,fig.width=7}
varImpPlot(rf.heart) # or importance(rf.heart)
```

* Chest pain still most important, but not by such
a wide margin.

## Number of Trees

* Bagging and random forests are not very sensitive to the number
of trees $B$.
* Can try different values and make sure $B$ large enough that
error has stabilized.

\vspace*{-.4in}

\begin{center}
\includegraphics[height=2.3in]{Fig8_8.pdf}
\end{center}

\vspace*{-.4in}

\small

* Text, Fig. 8.8. Not sure what "Test" error is based on,
but assume it is from a hold-out validation sample.
OOB error is lower than test error; maybe trees
are based on more data (no hold-out)?