---
title: "Assignment 4"
author: "Name 1, Name 2, Name 3"
date: "2019-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Fill in your computations and answers to the assignment questions in this
RMarkdown document. When you are finished, click the "Knit" button on RStudio
to render an HTML document. You can then use your browser or tool of choice
to convert the HTML document to a PDF file.

This assignment is to be handed in through canvas on Monday Oct 7 at 11:00pm.
(Note that this due date is different from the due date given on the canvas
Admin page.) This is a group assignment.
You must join a group on canvas even if you want to work alone. Please upload one PDF file
with your solutions per group. 


## Question 1 (Chapter 6, #8, parts (a)-(e), 10 marks)


(a) (1 mark)
(Note: You should set your random seed, for reproducibility.)

(b) (1 mark)


(c) (3 marks)
For the "best model obtained", you should 
use one that is parsimonious and close to
the consensus best according tht the three
selection criteria.

You don't **have** to create a data frame. 
`regsubsets()` can take a design matrix and
response vector, just like `lm.fit()` and 
`glmnet()`. If you do decide to create a data frame,
the following hint may be of use:
```{r}
library(leaps)
pmax <- 10
Xmat <- matrix(NA,nrow=n,ncol=pmax)
for(i in 1:pmax) {
  Xmat[,i] <- X^i
}
colnames(Xmat) <- paste0("X.",1:pmax)
dat <- data.frame(Y,Xmat)
```


(d) (2 marks) 


(e) (3 marks)

```{r}
library(glmnet)
lambdas <- 10^{seq(from=-2,to=5,length=100)}
cv.lafit <- cv.glmnet(Xmat,Y,alpha=1,lambda=lambdas) 
plot(cv.lafit)
la.best.lam <- cv.lafit$lambda.1se
la.best.lam
la.best <- glmnet(Xmat,Y,alpha=1,lambda=la.best.lam)
coef(la.best)
```


## Question 2 (Ch6, #9, 12 marks)

(a) (0 marks)
To make everyone's results comparable, please
select your test set with the following.
(Note that we scale all columns, including the response.)

```{r}
library(ISLR)
data(College)
library(dplyr)
College <- mutate(College,Private = as.numeric(Private=="Yes"))
College <- data.frame(lapply(College,scale))
dim(College) # 777 rows, use 111 as test
set.seed(1)
testset <- sample(1:777,size=111)
College.test <- College[testset,]
College.train <- College[-testset,]
```

(b) (2 marks)

(c) (2 marks)

(d) (2 marks)

(e) (2 marks)

(f) (2 marks)

(g) (2 marks)


## Question 3 (Ch7, #6, 8 marks)

(a) (5 marks)

(b) (3 marks)

