---
title: "Assignment 3"
author: "Name 1, Name 2, Name 3"
date: "2019-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Fill in your computations and answers to the assignment questions in this
RMarkdown document. When you are finished, click the "Knit" button on RStudio
to render an HTML document. You can then use your browser or tool of choice
to convert the HTML document to a PDF file.

This assignment is to be handed in through canvas on Monday Oct 7 at 11:00pm.
(Note that this due date is different from the due date given on the canvas
Admin page.) This is a group assignment.
You must join a group on canvas even if you want to work alone. Please upload one PDF file
with your solutions per group. 

## Question 1 (Chapter 3, #15)

(a)

(b)

(c)

(d)


## Question 2 (Chapter 4, #4)

(a)

(b) 

(c)

(d) 

(e)


## Question 3 (Chapter 4, #10 parts (a)-(h), 9 marks)

```{r}
library(ISLR)
data(Weekly)
head(Weekly)
```


(a)

(b) 

(c)

(d) 

(e) 

(f) 

(g) 

(h)  