---
title: "Week 2, Part b Exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* In these exercises we illustrate the bias-variance tradeoff 
by calculating the test error and expected test error for models
of different complexity.

## Test error 

```{r}
# parameters of the distribution of X
Xmin <- 10; Xmax <- 22
# paramters of the distribution of Y given X
beta0 <- 345; beta1 <- -72; beta2 <- 5; beta3 <- -.1; sd <- 4
# training and test set sample sizes
ntrain <- 30; ntest <- 10000 # not usually this large
# True f
f <- function(x) {
  return(beta0 + beta1*x + beta2*x^2 + beta3*x^3) # works if x is a vector
}
# data simulator
simdat <- function(n) { 
  x <- runif(n,min=Xmin,max=Xmax)
  y <- f(x) + rnorm(n,mean=0,sd=sd)
  return(data.frame(x=x,y=y))
}
set.seed(123)
traindat <- simdat(ntrain)
testdat <- simdat(ntest)

# Function to fit a polynomial model of a given degree on training data 
# and estimate the test error of this fitted model for a given test set.
testerr <- function(traindat,testdat,poly_degree) {
  fit <- lm(y~poly(x,poly_degree),data=traindat)
  fhat <- predict(fit,newdata=testdat)
  return( mean((testdat$y - fhat)^2) )
}

# Test error at x0=10 for different polynomial degrees
testerr(traindat,testdat,poly_degree=1)
testerr(traindat,testdat,poly_degree=2)
testerr(traindat,testdat,poly_degree=3)
testerr(traindat,testdat,poly_degree=4)
testerr(traindat,testdat,poly_degree=5)
testerr(traindat,testdat,poly_degree=6)
```

* **Exercise** Fit the model with best test error to the training data
and plot the results.

## Expected test error

* In the revised lecture slides, the expected test error was described as
either an average of conditional expected test errors, $Err(x_0)$
over the distribution of $X$, or the average of the test errors. 

* **Exercise** Run the following code. What do you notice about the 
behaviour of the expected test error compared to the test error.
(Hint: What was odd about the test error for polynomial degree 6?)

```{r}
# A version of testerr that takes only the polynomial degree as an argument
testerr2 <- function(poly_degree){
  traindat <- simdat(ntrain)
  testdat <- simdat(ntest)
  return(testerr(traindat,testdat,poly_degree))
}

exptesterr <- function(poly_degree,NREPS = 1000) {
  tes <- replicate(NREPS, testerr2(poly_degree) )
  return(mean(tes))
}

set.seed(123)
exptesterr(1)
exptesterr(2)
exptesterr(3)
exptesterr(4)
exptesterr(5)
exptesterr(6)
```

