---
title: "Week 3 Exercises"
author: "Brad McNeney"
date: '2018-09-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will do some model selection on a dataset called
`Boston` that contains house prices (`medv`, median house prices in \$1000) and characteristics
of neighborhoods in Boston.
We will fit models with interaction terms 
and will use a form of backward selection that 
tries to remove blocks of terms involving one of the
13 predictors of `medv`.

```{r}
library(MASS)
data(Boston)
head(Boston)
```


Fit a model that includes all possible predictors of 
`medv`.

```{r}
bfit <- lm(medv ~ ., data=Boston)
round(summary(bfit)$coefficients,4)
```

1. If you were to perform backward selection starting from the above
model, using p-values as the criterion for dropping terms, and
a threshold of $p > 0.05$ for stopping, would you drop any
terms? Which 
term would you drop first?

2. If you modify the formula from `medv ~ .` to  `medv ~ .*.` you 
will fit a model that includes all possible pair-wise interactions.
Fit this model and examine the coefficients summary. Which model
term would you drop first?

3. The formula `medv ~ .*. - zn:.` will remove from 
the model in (2)  the `zn` main effect and
all interaction terms that involve `zn`.
Fit a reduced model without `zn` terms and print the 
coefficients summary table. Use the `anova()` function to
perform a multiple-partial F-test for removing all `zn` terms.
On the basis of the multiple-partial F-test, would you simultaneously remove all
`zn` terms from the model?


4. Now remove from the model in (3) all terms involving `indus`. 
Would you remove these terms? 
