---
title: "Week 4 Exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will perform multiple regression on the `gapminder`
data to see which variables predict life expectancy
in 1952 for countries from Asia, Europe and Africa.

We first perform some data manipulations to (i) add log-10 GDP, (ii) subset
to the 1952 data, and (iii) remove year.

```{r}
library(gapminder)
library(dplyr)
data(gapminder)
gapminder <-
  gapminder %>% mutate(lgdpPercap = log10(gdpPercap),
                       lPop = log10(pop))
gm1952 <- gapminder %>% filter(year==1952) %>%
  filter(continent=="Africa" | continent=="Europe" | continent=="Asia") %>%
  select(-year,-gdpPercap,-pop)
```


1.
Fit a model for life expectancy
that includes your log-GDP and log-population
variables and interactions between these variables
and continent. 

2. Simultaneously
test the null hypothesis that log-population 
and its interactions with continent
do not modify the effect of log-population
on life expectancy. Report the conclusion of
your test using a 5% significance level.

3. In light of part (3) we drop population from the model
and use the reduced model.
Write the fitted model, $\hat{f}(X)$, for 
life expectency for countries in Europe.
Interpret the slope of this equation.

4. Use `ggplot()` to plot the residuals
_versus_ fitted values. Comment on the plot.


5. Use the `qqPlot()` function from `car` to do a Q-Q plot 
of the Studentized residuals. Comment on the plot. What
does you plot suggest?

6. (I am doing this one for you. Please look at the 
code below and make sure you understand its purpose.)
Calculate the Studentized residuals,
leverages (hats) and Cook's distances. Use
`filter()` once to pick out any points that are serious outliers,
have very high leverage or are highly influential.

In the following, I'm using an lm object called `gfitRed` that contains
my final fitted model.


```
gmAug <- data.frame(gm1952,studRes = rstudent(gfitRed), hats = hatvalues(gfitRed),
cooks = cooks.distance(gfitRed))
filter(gmAug, abs(studRes)>3) # None
filter(gmAug, hats>3*6/115) # Bosnia and Herzegovina (low GDP), Kuwait (high GDP)
filter(gmAug, cooks > 1) # None.
# Now sort or View(gmAug)
```