---
title: "Week 12 Exercises"
author: "Brad McNeney"
date: '2017-11-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will perform PCA on the
the `decathlon` data set from the `FactoMineR` package.
There is a short quiz on Canvas named
`Week 12 Tutorial Quiz` for you to fill out after doing these questions.


```{r}
library(FactoMineR) #install.packages("FactoMineR")
data(decathlon)
head(decathlon)
library(dplyr)
# Data processing strips off row names. Save for later.
rnames <- rownames(decathlon)
# Extract Olympics competition
rnames <- rnames[decathlon$Competition=="OlympicG"]
decathlon <- filter(decathlon,Competition=="OlympicG") 
# Extract data on the 10 events.
decathlon <- decathlon[,1:10] # Extract data on the 10 events.
# In most events, a high score is good, but the opposite is true for running.
# Change the running to scores where high is good by subtracting the times from
# the maximum time
diffmax <- function(x) { max(x)-x }
decathlon <- mutate(decathlon,
                    `100m` = diffmax(`100m`),
                    `400m` = diffmax(`400m`),
                    `110m.hurdle` = diffmax(`110m.hurdle`),
                    `1500m` = diffmax(`1500m`))
rownames(decathlon) <- rnames
```

(1) Perform PCA on the data without centering or scaling. Interpret the first PC.


(2) Perform PCA on the centered and scaled data. Interpret the first PC.

(3) Perform a scree plot for the PCA of part (2). 
Is there a clear "elbow" in the plot?

(4) Perform a biplot of the PCA in part (2). Are there
any obvious clusters of individuals?