---
title: "Week 8 Exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the file `week8Util.R` are two R functions 
based on the Chapter 7 notes. We will use
the `Boston` data set from the `MASS` package.

```{r}
source("week8Util.R")
library(MASS)
data(Boston)
```


1. Read and understand the code of the function
`cv.lm()` in the `week8Util.R` source file.
This is a generalization 
of the `cv()` function used in the ch07part2 notes (page 16)
to perform cross-validation for a least-squares regression.
Cut-and-paste the following code to test out `cv.lm()`.

```{r}
library(ISLR)
data(Wage)
library(splines)
k<-10; nDf <- 10; seed <- 1; cvErrs <- rep(NA,nDf) 
for(df in 1:nDf){ # loop over df
  cvErrs[df] <- cv.lm(wage ~ ns(age,df),Wage,k,seed)$meanErr
}
# The following should look like the plot on p. 17 of ch07part2.pdf
plot(1:nDf,cvErrs,type="b",ylab="MSE",xlab="DF for natural spline") 
```

2. Read and understand the following code that 
loops over df, computes the CV estimate and 
its standard error, and plots the results.
The functions `plot.cv.lm()` and `plotfit()` are in the 
`week8Util.R` source file.

```{r}
cvDat <- matrix(NA,nrow=nDf,ncol=4)
for(df in 1:nDf) {
  res <- cv.lm(wage ~ ns(age,df),Wage,k,seed)
  merr <- res$meanErr; serr <- res$sdErr
  cvDat[df,] <- c(df,merr,merr-serr,merr+serr)
}
colnames(cvDat) <- c("df","meanErr","lwr","upr")
#
# Plot the results with 1-SE error bands
#
library(ggplot2)
dfs <- plot.cv.lm(cvDat)
#
# Fit the best model and plot, with error bands. Use
# the plotfit() function from the chapter 7 notes.
#
fit <- lm(wage ~ ns(age,dfs$df.1se),data=Wage)
newdat <- data.frame(age=seq(from=min(Wage$age),to=max(Wage$age),length=100))
plotfitWage(fit,Wage,newdat)
```

3. Load the `Boston` data set, use the variable `dis`
(weighted mean of distances to five Boston employment centres) 
as a response 
and `nox` (nitrogen oxides concentration, in ppm) as
a predictor
in the following questions.
    (a) Use `poly()` to fit polynomials of degree
    1 to 10. Select the degree by cross-validation 
    using the 1-SE rule. (You will probably want to 
    modify the code in part 2 for this.)
    Plot the fitted model for
    the selected degree.
    (b) Repeat (a) using `ns()` to fit natural cubic splines with
    df 1 to 10. 
    (c) Fit a smoothing spline to these data and 
    plot the fitted model with the optimal df.
    (d) Comment on the differences between the fit of the
    selected models in (a)-(c)
    