---
title: "Week 8 Exercises"
author: "Brad McNeney"
date: '2018-10-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use a data set collected on a population at risk 
for type 2 diabetes. 
The data is from a study of type 2 diabetes in 
the Pima first nation. A description of the data
is in the text file `pima-diabetesDescr.txt` 
in the `Data` folder on GitHub.
The data on individual
characteristics like blood glucose and insulin 
were collected for 
predicting type 2 diabetes status (variable `Outcome`)
but we will use the variables other than `Outcome`
to predict `Glucose`.

```{r}
Diab <- read.csv("../Data/pima-diabetes.csv")
head(Diab,n=3)
summary(Diab)
```

There are many zeros in the data that are implausible 
and probably represent missing values. 
You should use your own judgment on recoding missing 
values. I am going to recode all the zeros for
`Glucose`, `BloodPressure`, `BMI` and `SkinThickness` 
as missing (`NA`), and then remove subjects
with missing data from the dataset.

```{r}
Diab$Glucose[Diab$Glucose==0] <- NA
Diab$BloodPressure[Diab$BloodPressure==0] <- NA
Diab$BMI[Diab$BMI==0] <- NA
Diab$SkinThickness[Diab$SkinThickness==0] <- NA
Diab <- na.omit(Diab)
```


1. Remove `Outcome` (diabetes status) from the data.
All remaining variables except `Glucose` will 
be used to predict `Glucose`. 
Compute the mean and variance of each predictor.


2. Next compute principal components with 
`prcomp()` twice. In your first call to `prcomp()`,
use the default options. In your second, set
`center=TRUE` (the default) and `scale=TRUE`
to centre and scale the data matrix before computing PCs.
Plot the two outputs of `prcomp()` to visualize the variance
attributable to each PC. For each output,
does the variance drop off
quickly with PC number, or does it decrease gradually?
To what do you attribute the difference between plots?

3. Centre and scale all the predictor variables to 
have mean 0 and SD 1. Hint: `data.frame(lapply(X,fun))` will
apply function `fun` to all columns of data frame `X` 
and will return the results as a data frame.


4. Set the random number seed to `1122` and
then use `pcr()` from the `pls` package to perform 
PCR using `Glucose` as the response 
and the centred and scaled predictors from part (3)
as predictors. Based on the cross-validation,
how many PCs would you choose?
Inspect the PC loadings (the `loadings` component
of the output of `pcr()`). Can you interpret 
these loadings to understand what 
factors influence blood glucose?

