---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 3, Part 4: Linear Regression vs K-Nearest Neighbors'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
library(ggplot2)
uu <- url("http://faculty.marshall.usc.edu/gareth-james/ISL/Advertising.csv")
advert <- read.csv(uu,row.names=1)
afit2<-lm(sales ~ TV + radio,data=advert)
library(dplyr)
advert <- mutate(advert,cTV = TV - mean(TV),
         cRadio = radio - mean(radio))
afitC <- lm(sales ~ cTV*cRadio+I(cTV^2) +I(cTV^3)+I(cTV^4),
            data=advert)
```


## Parametric vs Non-Parametric

* Linear regression is parametric.
    + We assume a functional (parametrized) form for $f$, and 
    then fitting $f$ amounts to fitting parameters.
* Non-parametric regression does not assume a parametric
form for $f$.
    + More flexible.
    + A simple example is $K$-nearest neighbors (KNN)
    regression.
    

## KNN Regression

* Define a neighborhood size $K$.
* For each $x_i$, take $\hat{f}(x_i)$
to be the average of the $y_j$'s for 
$x_j$'s in the neighborhood of $x_i$.

## Example Neighborhood

\scriptsize

```{r}
Xdat <- advert[,c("TV","radio")]
dm <- as.matrix(dist(Xdat))
dm[1:3,1:3]
dd <- dm[,1] # distances from first x
nbrThresh <- sort(dd)[9] # find 9th smallest distance
nbrThresh
nn <- (dd <= nbrThresh)
nn[1:3]
```


## 

* Encapsulate computation in a function.


```{r}
nbhd<-function(index,dat,K){
  dd <- as.matrix(dist(dat))[,index] 
  nbrThresh <- sort(dd)[K]
  return(factor(dd <= nbrThresh))
}
advert <- mutate(advert,nbHood = nbhd(1,Xdat,K=9))
```

##

\scriptsize

```{r,fig.height=1.5,fig.width=6}
advert[1,]
ggplot(advert,aes(x=TV,y=radio,color=nbHood)) + geom_point()
```

## KNN Prediction for First City


```{r}
advert[1,]
with(advert,mean(sales[nbHood==TRUE]))
```

## KNN Predictions for Advertising Data


```{r}
n <- nrow(advert)
K <- 9
KNNpred <- rep(NA,n)
for(i in 1:n) {
  advert <- mutate(advert,nbHood=nbhd(i,Xdat,K))
  KNNpred[i] <- with(advert,mean(sales[nbHood==TRUE]))
}
```

##

\scriptsize

```{r,fig.height=2.5,fig.width=4}
mutate(advert,fitted = KNNpred, residuals=sales-KNNpred) %>%
  ggplot(aes(x=fitted,y=residuals)) + geom_point() + 
  geom_smooth()
```

## Example from Text

* Plots of $\hat{f}(X)$ using KNN on 64 observations with
$K=1$ (left panel) and $K=9$ (right panel).

\begin{center}
\includegraphics[height=1.5in]{Fig3_16.pdf}
\end{center}

* For $K=1$ the KNN interpolates and for
$K=9$ it smooths.
    + Which is best? The one that gives 
    the best test set error rates 
    + We will discuss methods for esimating
    the test set error rate, but for
    now we simply break the advertising
    data into a training and test set.
    

## Test Set Predictions from KNN

* For a prediction point $x_0$, find the 
neighborhood $\mathcal{N}_0$ of the $K$
closest points in the training set, and take
$$
\hat{f}(x_0) = \frac{1}{K} \sum_{x_i \in \mathcal{N}_0} y_i
$$


## KNN on Advertising Data

\scriptsize

```{r}
library(caret) # install.packages("caret") if not already done
Y <- advert$sales
trainset <- (1:(.8*n))
trainX <- Xdat[trainset,]
trainY <- Y[trainset]
testset <- ((.8*n+1):n)
testX <- Xdat[testset,]
testY <- Y[testset]
maxK <- 10
testMSE <- rep(NA,maxK)
for(k in 1:maxK) {
  fit <- knnreg(trainX, trainY, k)
  testMSE[k] <- sum((testY - predict(fit, testX))^2)
}
```

##

\scriptsize

```{r,fig.height=2.5,fig.width=4}
data.frame(MSE = testMSE, K=1:maxK) %>%
  ggplot(aes(x=K,y=MSE)) + geom_point()
```

\normalsize

* According to the current test set, $K=2$ 
gives the best test set MSE.


## KNN with $K=2$ on Advertising Data

\scriptsize

```{r,fig.height=2.5,fig.width=4}
K <- 2
for(i in 1:n) {
  advert <- mutate(advert,nbHood=nbhd(i,Xdat,K))
  KNNpred[i] <- with(advert,mean(sales[nbHood==TRUE]))
}
mutate(advert,fitted=KNNpred,residuals=sales-KNNpred) %>%
  ggplot(aes(x=fitted,y=residuals)) + geom_point() + geom_smooth()
```

## Comparison of KNN to Linear Regression

* The text compares KNN to linear regression 
(with main effects) under different relationships
(linear or non-linear) and different numbers
of predictors $p$.


## Linear $f$, $p=1$.

* When the true $f$ is linear, linear regression 
MSE (dotted line) is slightly better than KNN 
MSE (green).

\begin{center}
\includegraphics[height=2.5in]{Fig3_18.pdf}
\end{center}

## Non-Linear $f$, $p=1$

\begin{center}
\includegraphics[height=3in]{Fig3_19.pdf}
\end{center}


## Non-Linear $f$, Varying $p$

* KNN is better for small $p$ but worse for 
large $p$

\begin{center}
\includegraphics[height=1.5in]{Fig3_20.pdf}
\end{center}

* When $p=20$, for example, the "nearest" neighbors
are not very near, and so do a poor job of 
predicting $f$.
    + This phenomenon is known as the 
    "curse of dimensionality".
