---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 7, Part 1: Simple Extensions of the Linear Model'
author: "Brad McNeney"
date: '2017-10-24'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3.5,fig.width=5)
```

## Polynomial Regression

* We have used polynomial regression before.
* This was the standard way to extend the linear model.
* The text recommends against a degree of more than 4.
* To avoid collinearity we can use the covariates
returned by the `poly()` function (more on these functions
in the next set of notes).

## Example: Wage Data

* Predict wages as a function of age.

\scriptsize

```{r}
library(ISLR)
data(Wage)
head(Wage,n=3)
```

##

\scriptsize

```{r}
library(ggplot2)
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=0.1) +
  geom_smooth(formula=formula(wage ~ poly(age,4)))
```

## DIY Confidence Bands

* Use `predict()` to get the predictions and plot CI bands with 
`geom_ribbon()`

* First the predictions and point-wise CIs.

\scriptsize

```{r}
wfit <- lm(wage ~ poly(age,4),data=Wage)
newWage <- data.frame(age = sort(unique(Wage$age)))
wpred <- data.frame(newWage,predict(wfit,newdata=newWage,interval="confidence"))
head(wpred,n=3)
```

##

\scriptsize

```{r}
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=0.1) +
  geom_ribbon(aes(x=age,y=fit,ymin=lwr,ymax=upr),
              data=wpred,fill="blue",alpha=.2) + 
  geom_line(aes(y=fit),data=wpred,color="blue")
```

## Classification Example: Wage $>$ 250K

\scriptsize

```{r}
library(dplyr)
Wage <- mutate(Wage,wage250 = as.numeric(wage>250))
wfit <- glm(wage250 ~ poly(age,4),data=Wage,family=binomial())
wpred <- data.frame(newWage,
          predict(wfit,newdata=newWage,se.fit=TRUE)) # predicted logits
expit <- function(x) { exp(x)/(1+exp(x)) }
wpred <- mutate(wpred,
                lwr=expit(fit-2*se.fit),upr=expit(fit+2*se.fit),
                fit = expit(fit))
# Don't go below 0 or above 1
head(wpred,n=3)
```


##

\scriptsize

```{r}
ggplot(wpred,aes(x=age,y=fit)) + geom_rug() +
  geom_ribbon(aes(ymin=lwr,ymax=upr),fill="blue",alpha=.2,limits=c(0,.2)) + 
  geom_line(aes(y=fit),data=wpred,color="blue") 
```

##

\scriptsize

```{r}
wpred <- mutate(wpred,upr = pmin(upr,0.2))
ggplot(wpred,aes(x=age,y=fit)) + geom_rug() +
  geom_ribbon(aes(ymin=lwr,ymax=upr),fill="blue",alpha=.2,limits=c(0,.2)) + 
  geom_line(aes(y=fit),data=wpred,color="blue") 
```

## Step Functions

* In Epidemiology it is common to discretize age and treat
as a categorical variable.
    + Categorical variables are coded as dummy variables for regression.
* The regression will fit a separate mean for each category, which 
is more flexible than, say, linear in age.

\scriptsize

```{r}
with(Wage,range(age))
agebreaks <- c(15,20,25,30,35,40,45,50,55,60,65,70,75,80)
wfit <- lm(wage ~ cut(age,breaks=agebreaks),data=Wage)
newWage <- data.frame(age = sort(unique(Wage$age)))
wpred <- data.frame(newWage,predict(wfit,newdata=newWage,interval="confidence"))
head(wpred,n=3)
```

##

\scriptsize

```{r}
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=0.1) +
  geom_ribbon(aes(x=age,y=fit,ymin=lwr,ymax=upr),
              data=wpred,fill="blue",alpha=.2) + 
  geom_line(aes(y=fit),data=wpred,color="blue")
```

## Classification Example: Wage $>$ 250K

* No top-earners less than about 25 or older than about 
65, so need to re-do the age breaks.

\scriptsize

```{r}
library(dplyr)
agebreaks <- c(15,30,35,40,45,50,55,60,80)
wfit <- glm(wage250 ~ cut(age,agebreaks),data=Wage,family=binomial())
wpred <- data.frame(newWage,
             predict(wfit,newdata=newWage,se.fit=TRUE)) # predicted logits
expit <- function(x) { exp(x)/(1+exp(x)) }
wpred <- mutate(wpred,
                lwr=expit(fit-2*se.fit),upr=expit(fit+2*se.fit),
                fit = expit(fit))
head(wpred,n=3)
```

##

\scriptsize

```{r}
ggplot(wpred,aes(x=age,y=fit)) + geom_rug() +
  geom_ribbon(aes(ymin=lwr,ymax=upr),fill="blue",alpha=.2,limits=c(0,.2)) + 
  geom_line(aes(y=fit),data=wpred,color="blue") 
```