---
title: "Stat 652 Project Guidelines"
author: "Brad McNeney"
date: '2018-11-14'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data 

The data are from the Canadian Community Health Survey (CCHS) --
Healthy Aging module. Documentation for the study
is available in the Documentation folder. Please read
Sections 1 and 2 of the User Guide document 
(`CCHS_HA_User_Guide.pdf`). The project will have
two parts.

### Part I: Health Utility Indices and Cognitive Health Status

The file `HUI.csv` contains a training dataset comprised
of 8 health-utility-index (HUI)
variables. The 8 are a subset of the 13 variables described
on pages 45-54 of the file `CCHS_HA_Derived_variables.pdf`.
The first variable, `HUIDCOG` is a categorical variable
with six levels of cognitive ability. See the Derived variables
document for a description of the levels.

```{r}
hui <- read.csv("hui.csv")
dim(hui)
summary(hui)
```

Use the other HUIs to predict `HUIDGOC`. I have a 
hold-out test dataset on which you will eventually evaluate
your predictions.

### Part II: Predicting a HUI Score with Other Variables

In this part of the project you will 
use a larger (wider) dataset called `HStrain` 
that consists of 591 variables measured on 
10000 subjects.
You will find these data in the file `HStrain.csv.zip` in the 
Data directory; unzip it and 
use the R function `read.csv()` to read it into R.
I have removed all of the subjects 
with `NOT STATED` entries in any of the 591 variables.

In the dataset, the categorical
HUI variables have been removed and
replaced by a single
score called `HUIDHSI`. The description of `HUIDHSI`
is as follows:

*This derived variable is a Health Utilities Index which provides a description of an individual’s overall functional health, based
on eight attributes: vision, hearing, speech, ambulation (ability to get around), dexterity (use of hands and fingers), emotion
(feelings), cognition (memory and thinking) and pain (in HUP module). The version of the index used in CCHS is adapted from
the HUI Mark 3 (HUI3). The index is designed to produce an overall health utility score. This multi-attribute utility index
produces a score ranging from 1.000 (perfect health), through 0.000 (health status equal to death) to -0.360 (health status
worse than death).*

Your task is to predict `HUIDHSI` with the
other 590 variables in the `HStrain` dataset.
You may work with `HUIDHIS` as-is (a quantitative
variable), **or** break it into a binary variable
that has value 0 if `HUIDHIS` is less than the 
median value of 0.905, and 1 otherwise.
I have a 
hold-out test dataset on which you will eventually evaluate
your predictions. 

## Project Length and Scope

Your report should be no more than 5 pages long, plus
references. You must also include an Appendix of R code that
can be used to reproduce the analyses refered to in
the report. There is no page limit for the Appendix, but
please use judgement about what to include. Too long
and it is not likely to be read.
You are encouraged to 
try several prediction methods, and can compare 
these methods, but your report should focus on 
one method in particular. 

## Grading Criteria

The criteria for the report are as follows.

### Report (25 marks)

The report should have the following sections

1. Introduction (brief)
2. Data (brief)
3. Methods
4. Results
5. Conclusions and Discussion  
    
The reports will be judged on the following criteria.

* Content (17): 
The content should be clear, accurate, complete and at the 
level of students in Stat 452. In the Methods you should provide a brief
description of any statistical methods you use. Please
restrict yourself to methods that were covered
in class. You can mention methods not covered as
areas of future work. 
Methods you considered 
but were not the focus of your report should be briefly mentioned 
here too.
In Results you should summarize and interpret the fitted model. 
Though the primary goal is prediction, your insights into the 
data-generating process are important. Refer to the Appendix for
the code that implements your prediction equation.
In the Conclusions and Discussion present your conclusions, 
discuss short-comings of your approach, and, optionally,
ideas for further work.
* Organization (3): Though the report is structured, you should
present your ideas logically within each section.
* References (3): You must properly attribute the ideas and 
work of others.
* Grammar and spelling (2): Please proof-read your report.

### Code (15 marks)

The code in your Appendix
should be correct, readable and should knit without
errors. A secondary consideration is efficiency. 
The Appendix will be judged on the following criteria.

* Software Details (2): List the version of R you are using
and the names of all packages used in your analysis **at the beginning**
of the Appendix. Please also provide an estimate of the time
it will take to knit the code if more than about 2 minutes.
* Correctness (5): There should be no errors in data
processing, function calls, etc.
* Readability (5): The steps of your analysis should 
be clearly layed out and it should be easy for the reader
to find the final prediction equation/method.
* Efficiency (3): Please take steps to avoid computational
inefficiencies, such as loops and excessive copying of 
large R objects.
