---
title: "Stat 652 Project Guidelines"
author: "Brad McNeney"
date: '2019-10-16'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data 

The data are on flights from three New York City airports 
in 2013, from the `nycflights13` package. 
Data were combined from four datasets from this package:

* `flights`
* `weather`
* `airports`, and
* `planes`

Please read about the variables in each dataset by typing
`help(datasetname)` from the R console. 

```{r}
library(tidyverse)
library(nycflights13)
#help(flights)
#help(weather)
#help(airports)
#help(planes)
fltrain <- read_csv("fltrain.csv.gz")

# when the test data are made available:
# fltest <- read_csv("fltest.csv.gz")
```

Your task is to build a prediction model for 
departure delays. I have held out 1336776 observations
as a test dataset that I will release a few days 
before the project is due.

## Project Length and Scope

Your report should be no more than 5 pages long, plus
references. You must also include an Appendix of R code that
can be used to reproduce the analyses refered to in
the report. There is no page limit for the Appendix, but
please use judgement about what to include. Too long
and it is not likely to be read.
You are encouraged to 
try several prediction methods, and can compare 
these methods, but your report should focus on 
one method in particular. 
You **must** use methods discussed in class.

## Grading Criteria

The criteria for the report are as follows.

### Report (25 marks)

The report should have the following sections

1. Introduction (brief)
2. Data (brief)
3. Methods
4. Results
5. Conclusions and Discussion  
    
The reports will be judged on the following criteria.

* Content (20): 
The content should be clear, accurate, complete and at the 
level of students in Stat 452. In the Methods you should provide a brief
description of any statistical methods you use. Please
restrict yourself to methods that were covered
in class. You can mention methods not covered as
areas of future work. 
Methods you considered 
but were not the focus of your report should be briefly mentioned 
here too.
In Results you should summarize and interpret the fitted model. 
Though the primary goal is prediction, your insights into the 
data-generating process are important. Refer to the Appendix for
the code that implements your prediction equation.
In the Conclusions and Discussion present your conclusions, 
discuss short-comings of your approach, and, optionally,
ideas for further work.
* Organization (3): Though the report is structured, you should
present your ideas logically within each section.
* Grammar and spelling (2): Please proof-read your report.

### Code (15 marks)

The code in your Appendix
should look correct and be readable. 
Given the size of the dataset, you do
not need to provide run-able code for
all analyses. Use `{r, eval=FALSE}` in your computationally-intensive code chunks
to prevent them from running.
The Appendix will be judged on the following criteria.

* Software Details (2): List the version of R you are using
and the names of all packages used in your analysis **at the beginning**
of the Appendix. Please also provide an estimate of the time
it will take to knit the code if more than about 2 minutes.
* Correctness (5): There should be no errors in data
processing, function calls, etc.
* Readability (5): The steps of your analysis should 
be clearly layed out and it should be easy for the reader
to find the final prediction equation/method.
* Efficiency (3): Please take steps to avoid computational
inefficiencies, such as loops and excessive copying of 
large R objects.
