---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 10, part 2: Cluster Analysis'
author: "Brad McNeney"
date: '2018-11-14'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Clustering

* Clustering is a set of techniques for finding *subgroups*, or
clusters, in a data set.
* A good set of clusters is a partition of the data such that
observations
are *similar* within clusters 
and *dissimilar* between clusters.
* Example: Suppose we have $n$ observations of $p$ features for cancer
patients. 
    + We can cluster to look for cancer sub-types.
    + However, these clusters will be highly dependent on how we measure similarity/dissimilarity.  
    + Plus, we don't *know* that clusters correspond to cancer sub-types.


## Clustering as Dimension Reduction

* We can view clustering as an exploratory method to understand
possible structure in our data.
* Similar in spirit to PCA, but a different mechanism
    +  PCA finds a low-dimensional representation of observations explaining a good proportion of the variance
    + Clustering looks to find homogeneous sub-groups.

## Clustering Methods

* There are many.
* We focus on $K$-means/medoids and hierarchical clustering,
and assume all features are quantitative.
* In $K$-means we partition into a pre-specified number of
clusters.
* In hierarchical clustering we successively group observations.
    + Represent the nested partitions as a tree-like
    structure called a *dendrogram*.

## K-means Clustering

* Choose the desired number of clusters, $K.$  The $K$-means clustering algorithm will assign each observation to *exactly one* of the $K$ clusters.
* Let $C_k$ be the set of indices for observations assigned to cluster
$k$, $k=1,\ldots,K$.
* $K$-means clustering chooses clusters to solve
$$\min_{C_1,\ldots,C_K}\sum_{k=1}^KW(C_k),$$
where $W(C_k)$ is a measure of the amount by which observations within cluster $C_k$ differ from one another.


## K-means Clustering: Choice of $W(C_k)$

* For continuous variables, the standard
choice for $W(C_k)$ is the squared Euclidean distance,
$$W(C_k)=\frac{1}{\vert C_k\vert}\sum_{i,i^\prime\in C_k}\sum_{j=1}^p(x_{ij}-x_{i^\prime j})^2$$
where $\vert C_k\vert$ denotes the number of observations in cluster $k$.
* It turns out that
$$\frac{1}{\vert C_k\vert}\sum_{i,i^\prime\in C_k}\sum_{j=1}^p(x_{ij}-x_{i^\prime j})^2=2\sum_{i\in C_k}\sum_{j=1}^p(x_{ij}-\bar{x}_{kj})^2$$
where the *cluster centroid* $\bar{x}_{kj}=\frac{1}{\vert C_k\vert}\sum_{i\in C_k}x_{ij}.$
* We can see where the name ``$K$-means'' comes from.

## K-means Clustering Algorithm

1. Randomly assign a number, from $1,\ldots,K$ to each of the 
observations.  They serve as the initial cluster assignments 
for the observations.
2. Iterate the following steps until the cluster assignments 
stop changing:
    (a) For each of the $K$ clusters, compute the cluster centroid.  
    The $k$th cluster centroid is the vector of the $p$ feature 
    means for the observations in the $k$th cluster for 
    continuous variables.
    (b) Assign each observation to the cluster whose centroid is 
    closest in terms of Euclidean distance.


## K-means Clustering Algorithm, Simulated Data Example

\begin{center}
\includegraphics[height=3in]{Fig10_6.pdf}
\end{center}

\scriptsize

* Text, Figure 10.6. 
Progress of the $K$-means algorithm with $K=3.$  



## K-means Clustering and Local Minima

* The algorithm is guaranteed to decrease the value of 
$\sum_kW(C_k)$ until no further improvement is possible, 
resulting in a *local* minimum.
* However, the local min depends on the random initialization.  
Hence, in practice we re-run the algorithm several 
times and keep the solution that achieves the lowest 
overall criterion value.
* See the video demo at http://www.youtube.com/watch?v=zaKjh2N8jN4.



## K-means Clustering with Different Initialization

\begin{center}
\includegraphics[height=2.5in]{Fig10_7.pdf}
\end{center}
\scriptsize
Text, Fig 10.7. $K$-means clustering performed six times on the same data with $K=3$ different random initializations.  Above each plot is the value of the objective function. Three different local optimums were found.
The overall minimum is 235.8.

## $K$ Means Clustering of the Iris Data

* We know there are three species of iris. Ignore the
species labels and do the clustering.

\scriptsize

```{r}
library(ggplot2)
data(iris) # help(iris)
head(iris) # plot sepal length vs width with labels
```

##

\scriptsize

```{r}
ggplot(iris,aes(x=Sepal.Width,y=Sepal.Length,color=Species)) + 
  geom_point()
```

## 

\scriptsize

```{r}
library(dplyr)
irisX <- iris %>% select(-Species) %>% scale()
kout3 <- kmeans(irisX,centers=3,nstart=10)
iris <- data.frame(iris,Clusters=factor(kout3$cluster))
with(iris,table(Species,Clusters))
```

##

\scriptsize

```{r}
ggplot(iris,
       aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Clusters)) +
  geom_point()
```

##


\scriptsize

```{r}
kout2 <- kmeans(irisX,centers=2,nstart=10)
iris$Clusters <- factor(kout2$cluster)
with(iris,table(Species,Clusters))
```

##

\scriptsize

```{r}
ggplot(iris,
       aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Clusters)) +
  geom_point()
```

## Scaling Variables

*  Whether or not to scale variables depends on the context,
but usually we will.

## Choosing $K$

* Like PCA, there is no "best" method for choosing $K$.
* Cross-validation is not an option because there is no outcome.
* For small $p$, can visualize the clusters, but this
becomes difficult as $p$ grows.
* The silhouette plot is a graphical approach.
    + Discussed after the PAM algorithm below.


## Sensitivity to Outliers

* Means and Euclidean distances are sensitive to outliers and 
$K$-means depends on both
    + Means are cluster centroids.
    + Criterion to minimize is a sum of squared Euclidean
    distances.
    
## $K$-Medoids Clustering

* An alternative to $K$-means is $K$-medoids clustering. 
* Cluster centres are medoids, which are observations
chosen to represent each cluster.
* There is flexibility in choosing the dissimilarity measure 
    + Can choose a more robust measure, such as $\ell_1$ (so-called
    Manhattan) distance.
* An implementation of $K$-medoids is the PAM (partitioning 
around medoids) algorithm.

## PAM on the Iris Data

\scriptsize

```{r}
library(cluster)
pout3 <- pam(irisX,k=3)
iris$Clusters <- factor(pout3$cluster)
with(iris,table(Species,Clusters))
```

##

\scriptsize

```{r}
ggplot(iris,
       aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Clusters)) +
  geom_point()
```

##

\scriptsize

```{r}
pout2 <- pam(irisX,k=2)
iris$Clusters <- factor(pout2$cluster)
with(iris,table(Species,Clusters))
```

##

\scriptsize

```{r}
ggplot(iris,
       aes(x=Sepal.Width,y=Sepal.Length,color=Species,shape=Clusters)) +
  geom_point()
```

## The Silhouette Plot

* For each observation we compute a measure of cluster certainty,
called the silhouette width, that takes values
in $[-1,1]$.
    + Values near 1 indicate certainty that the 
    observation is in the right cluster and 
    + Values near $-1$ indicate that
    the observation is in the wrong cluster.
* Formula for silhouette width is given in on the next slide.
* If the silhouette values in a cluster are, say,
below average, this suggests the cluster can be merged with another.

## Silhouette Widths

* For observation $i$ in cluster $k$, 
the silhouette width $s(i)$ is defined as follows.
* Let $a(i)$ be the average dissimilarity between $i$ and all
other observations in cluster $k$.
* Let $d(i,C)$ be the average dissimilarity between $i$ and all observations
in cluster $C$.
* Let $b(i) = \min_C d(i,C)$.
* Then 
$$s(i) = \frac{b(i)-a(i)}{\max(a(i),b(i))}$$
    + Note: $a(i)$ is a *dissimilarity* measure, and should be
    $\ll$ $b(i)$ if $i$ confidently in cluster $k$.

## Silhouette Plots for the Iris Data

\scriptsize

```{r}
sil3 <- silhouette(pout3)
plot(sil3,main=""); abline(v=mean(sil3[,"sil_width"]))
```

## 

\scriptsize

```{r}
sil2 <- silhouette(pout2)
plot(sil2,main=""); abline(v=mean(sil2[,"sil_width"]))
```

## 

\scriptsize

```{r}
sil5 <- silhouette(pam(irisX,k=5))
plot(sil5,main=""); abline(v=mean(sil5[,"sil_width"]))
```

