---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 10, part 1.5: Introduction to Multiple Correspondence Analysis'
author: "Brad McNeney"
date: '2018-11-14'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Multiple Correspondence Analysis (MCA)

* An exploratory analysis methodology for 
multivariate datasets with categorical variables.
* In basic form, it is PCA on dummy variables 
that represent the categorical variables.
* Illustrate with the health utilities index (HUI)
variables from the Canadian Community Health
Survey - Healthy Aging

## HUI Data

* Read from the Stat 652 Project folder.

\scriptsize

```{r}
hui <- read.csv("../../Project652/HUI.csv")
names(hui)
```

## Summaries

\scriptsize

```{r}
summary(hui)
```

## Remove records with missing values

* I will consider the response `NOT STATED` 
to be missing data. 
    + Remove subjects with any missing data
    
\scriptsize

```{r}
recode_ns <- function(x) {
  x[x=="NOT STATED"] <- NA
  x <- droplevels(x)
  x
}
for(i in 1:ncol(hui)) {
  hui[,i] <- recode_ns(hui[,i])
}
hui <- na.omit(hui)
dim(hui)
```

## `HUIDCOG`

* Cognitive function (our focus) with levels:

\small

1. Able to remember most things, think clearly and solve day to day problems
2. Able to remember most things, but have a little difficulty when trying to think and solve day to day problems
3. Somewhat forgetful, but able to think clearly and solve day to day problems
4. Somewhat forgetful, and have a little difficulty when trying to think or solve day to day problems
5. Very forgetful, and have great difficulty when trying to think or solve day to day problems
6. Unable to remember anything at all, and unable to think or solve day to day problems

##

\scriptsize

```{r}
levels(hui$HUIDCOG) <- as.character(1:6)
table(hui$HUIDCOG)
```

## Pairwise summaries

* Relationship between `HUIDCOG` and others

\scriptsize

```{r}
tt <- xtabs(~DHHGAGE+HUIDCOG,data=hui)
tt
```

## Pairwise summaries, cont.

* Age distributions for each cognitive level
    + Proportions of column variable for each 
    row (level of `HUIDCOG`).

\scriptsize

```{r}
round(prop.table(tt,margin=2),2)
```

## Pairwise summaries, cont.

* Relationship between `HUIDCOG` and `HUIDEX`.

\scriptsize

```{r}
tt <- xtabs(~HUIGDEX+HUIDCOG,data=hui)
tt
```

## Pairwise summaries, cont.

\scriptsize

```{r}
round(prop.table(tt,margin=2),2)
```

\normalsize

* And so on ...

## Correspondence Analysis (CA)

* Write the categorical variables as dummy variables
and do a biplot of the result; e.g., for the 
HUIs.

```{r,cache=TRUE}
library(dplyr)
hsub <- select(hui,starts_with("HUI"))
X <- model.matrix(~.,data=hsub)[,-1]
X <- scale(X)
pp <- prcomp(X)
# biplot(pp) -- too many points, too messy
```

##

```{r}
library(FactoMineR)
res.mca <- MCA(hsub)
```

##

\scriptsize

```{r}
plot(res.mca,invisible=c("ind"),cex=0.5)
```


