---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 6, Part 2: Shrinkage Methods'
author: "Brad McNeney"
date: '2018-10-12'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3.5,fig.width=5)
```


## Shrinkage Methods

* Fit a model that contains all $p$ predictors 
using a method that shrinks the coefficient 
estimates towards zero.

* This biases the estimates, but reduces variance.

* We will discuss two shrinkage methods, 
ridge regression and the lasso.

## Ridge Regression

* Penalize the criterion function, RSS, to favour
smaller coefficient values.
* The ridge regression criterion function is
$${\rm RSS} + \lambda \sum_{j=1}^p \beta_j^2$$
where $\lambda \geq 0$ is a tuning parameter. 
* The ridge regression estimator $\hat{\beta}^R_{\lambda}$
is the minimizer of this criterion.
* The penalty term has two components, 
the tuning parameter and the 
sum of squared coefficients.

## Tuning Parameter, $\lambda$

+ We do **not** penalize the intercept.
+ $\lambda=0$ gives least squares
+ $\lambda>0$ will lead to estimates of 
$\beta_1,\ldots,\beta_p$ that are "shrunken"
towards zero
+ To be practical, we need a method for choosing
    the tuning parameter.
    
## SS Coefficients, $\sum_{j=1}^p \beta_j^2$

* $\sum_{j=1}^p \beta_j^2$ is
the square of the length of the vector
$(\beta_1,\ldots,\beta_p)$ of non-intercept
coefficients.
* The length is the Euclidean or $\ell_2$ norm
of the vector.
    + Sometimes ridge regression is called
    $\ell_2$-penalized regression.

## Scaling Predictors

* The least squares solution is said to be 
scale invariant.
    + If we multiply
    a predictor $X_j$ by a constant $c$, the
    least squares solution $\hat{\beta}_j$ 
    is multiplied by $1/c$ so that
    $X_j \hat{\beta}_j$ doesn't change.
* The same is not true for ridge regression.
    + $X_j \beta^R_{\lambda,j}$ depends on the scale
    of $X_j$; e.g., on the units $X_j$ is
    measured in.
* We typically standardize each predictor
by subtracting its mean and dividing by its
sample SD.
    + Then the units of each $X_j$ don't matter.
* Aside: If we also standardize the response, then
it turns our the fitted intercept is zero.
    
## Application to Credit Data

\scriptsize

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Credit.csv")
Credit <- read.csv(uu,row.names=1)
head(Credit,n=3)
```

## Least Squares for Comparison

* Set up the design matrix and response 
ourselves and pass to the `lm.fit()`
function, which does the fitting for `lm()`.

\scriptsize

```{r}
Xfull <- model.matrix(Balance ~ ., data=Credit)
head(Xfull,n=3)
Y <- Credit$Balance
```

##

\scriptsize

```{r}
# Standardize predictors
predInds <- 2:ncol(Xfull) # exclude intercept
Xfull[,predInds] <- scale(Xfull[,predInds]) 
Y <- Credit$Balance
lsfit <- lm.fit(Xfull,Y)
lsfit$coefficients
```


## Ridge Regression 

* Find the ridge regression solution for each 
$\lambda$ on a grid.

\scriptsize

```{r}
library(glmnet) # install.packages("glmnet"), if necessary
Xfull <- Xfull[,-1] # glmnet will add intercept and scale X's
lambdas <- 10^{seq(from=-2,to=5,length=100)}
rrfit <- glmnet(Xfull,Y,alpha=0,lambda=lambdas)
round(cbind(coef(rrfit,s=lambdas[1]),coef(rrfit,s=lambdas[50])),4)
```

## Plotting coefficient "paths"

* It is customary to plot the estimated coefficients for
each $\lambda$ as a function of $\lambda$ or $\log\lambda$.

\scriptsize

```{r}
plot(rrfit,xvar="lambda")
```

##

\scriptsize

```{r}
plot(rrfit,xvar="lambda",ylim=c(-25,25))
```


## Bias-Variance Tradeoff

* Recall that the MSE is the variance plus bias
squared
* The least squares estimate of the regression
coefficients is unbiased and therefore
so are the predictions $X \hat{\beta}$.
* Penalizing introduces bias into the 
predictions, but reduces variance.
* Illustrated in the text for a "simulated dataset".
    + My guess is that they are evaluating
    variance and bias by simulating from a model.

##

\begin{center}
\includegraphics[height=5in]{Fig6_5.pdf}
\end{center}

* Figure 6.5 of the text.
The MSE is in purple, variance in green and 
squared bias in black.
    + Minimum MSE is at $\lambda$ of about 30.


## The Lasso

* A drawback of ridge regression is that it
does not *select* a subset of predictors.
    + The final model includes all $p$ coefficients,
    shrunken toward zero.
    + Not good for interpretation.
* An alternative called the lasso 
does model selection and shrinkage.
* The lasso replaces the 
$\ell_2$ penalty of ridge regression with 
an $\ell_1$ penalty; i.e., the lasso estimator
$\hat{\beta}^L$ minimizes the criterion
$$ {\rm RSS} + \lambda \sum_{j=1}^p |\beta_j|$$
* It turns out that the lasso can shrink estimates
to zero, and hence de-select predictors.
    + Variable-selected models are said to be sparse.

## The Lasso on the Credit Data

\scriptsize

```{r}
lafit <- glmnet(Xfull,Y,alpha=1,lambda=lambdas) # notice alpha=1
plot(lafit,xvar="lambda")
```

## 

\scriptsize

```{r}
plot(lafit,xvar="lambda",ylim=c(-25,25))
```

\normalsize

* After $\log \lambda>6$ or so all coefficients have
been set to zero.

## Equivalent Representation of Ridge and Lasso

* One can show that for a given $\lambda$ there 
is an $s$ such that the lasso solution $\hat{\beta}^L_{\lambda}$ is the solution to the 
constrained minimization of RSS subject to
$$\sum_{j=1}^p |\beta_j| \leq s$$
* Similarly, the ridge regression
solution $\hat{\beta}^R_{\lambda}$ 
is the solution to the 
constrained minimization of RSS subject to
$$\sum_{j=1}^p \beta_j^2 \leq s$$

##

\begin{center}
\includegraphics[height=4in]{Fig6_7.pdf}
\end{center}

* Figure 6.7 of the text.
The shaded regions are where the constraints
are satisfied for the lasso (left) and 
ridge regression (right). The 
contours are of the RSS. The lasso
solution zeroes out the $\beta_1$ coefficient.

## Selecting the Tuning Parameter

* We can select the $\lambda$ that minimizes
estimated test set error over a grid of 
$\lambda$ values.
    + Estimate test set error by cross-validation.
* Then fit the model with this best $\lambda$.
* Convenience function `cv.glmnet()` will
do most of the work for us.

## Credit Data Example, Ridge Regression

\scriptsize

```{r}
# Ridge regression
cv.rrfit <- cv.glmnet(Xfull,Y,alpha=0,lambda=lambdas)
plot(cv.rrfit)
```

##

\scriptsize

```{r}
plot(cv.rrfit,ylim=c(9000,12000))
```

## Error Bars

* The error bars are $\pm$ one SD of the MSE estimates
across the ten folds. 
* Hastie & co. (ESL, page 216)
advocate the "one-standard-error" rule: Use the most 
parsimonious model (largest $\lambda$)
whose error is no more than 
one SD above the error of the best model.
* Acknowledges that the MSEs are only estimates.
    + Rather _ad hoc_ rule though.
    
\scriptsize

```{r}
cv.rrfit$lambda.min; cv.rrfit$lambda.1se
```


## Fitted Model with Best $\lambda$

\scriptsize

```{r}
rr.best.lam <- cv.rrfit$lambda.1se
rr.best.lam
rr.best <- glmnet(Xfull,Y,alpha=0,lambda=rr.best.lam)
coef(rr.best)
```

##

* Now lasso

\scriptsize

```{r}
cv.lafit <- cv.glmnet(Xfull,Y,alpha=1,lambda=lambdas) 
plot(cv.lafit)
```

##

\scriptsize

```{r}
plot(cv.lafit,ylim=c(8000,13000))
```

##

\scriptsize

```{r}
la.best.lam <- cv.lafit$lambda.1se
la.best.lam
la.best <- glmnet(Xfull,Y,alpha=1,lambda=la.best.lam)
coef(la.best)
```

## Summary of Credit Data

* Ridge regression shrinks, but is not very interpretable.
* Lasso shrinks and selects variables.
    + The lasso solution is similar to the best model 
    found by model selection methods 
    (see week 6 exercises).

