---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 3, Part 3: Other Considerations in Multiple Linear Regression'
author: "Brad McNeney"
date: '2017-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
library(ggplot2)
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv")
advert <- read.csv(uu,row.names=1)
afit2<-lm(sales ~ TV + radio,data=advert)
```

## Topics

* Interaction and non-linear model terms
* Categorical variables as predictors
* Model diagnostics




## Example Data: Credit Card Balances

* Understand which variables are associated with 
credit card balance.

\scriptsize

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Credit.csv")
credit <- read.csv(uu,row.names=1)
head(credit)
```

## Software Note

* In R, categorical variables should be stored as **factors**.

# Interaction and Non-Linear Model Terms

## Interaction and Non-Linear Model Terms

* We have already seen non-linear model terms when we 
modelled the relationship between income and education
as a polynomial. 
* We now discuss interaction.


## Statistical Interaction

- Start with two explanatory variables income ($X_1$) and 
student status ($X_2$)
    - `StudentYes=1` if the person is a student and 0 otherwise.
- $X_2$ is said to modify the effect of $X_1$ on $Y$ if the regression 
slope of the regression 
of $Y$ on $X_1$ differs in the $X_2=0$ and $X_2=1$ sub-groups. 
    - If we stratify the analysis by student status and find different 
effects of income in the two groups, there is statistical interaction 
between income and student status.

## Model for Stratification by Student Status

\small

- $f(X) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1X_2$
where 
    - $X_1$ is income
    - $X_2$ is student status (1 is student, 0 is not)
    - $X_1 \times X_2$ is the statistical interaction between 
    income and student status.
    - $\beta_1$, $\beta_2$, and $\beta_3$ are the corresponding regression coefficients.
- This model allows separate lines for the two values of student
status.
    - student status = 0 model: intercept $\beta_0$ and slope $\beta_1$
    - student status = 1 model: intercept $\beta_0+\beta_2$ and slope $\beta_1+\beta_3$
    - Interpret $\beta_3$ as the difference between slopes. 
    - If $\beta_3=0$, then student status does not modify effect of income on balance.
    - In practice, we test the hypothesis $H_0:\beta_3=0$. 

## Fitted Model

\scriptsize

```{r}
cfit <- lm(Balance ~ Income*Student,data=credit)
summary(cfit)$coefficients
```

\small

- The $t$-test of the hypothesis $H_0:\beta_3=0$ does not reject at any of the standard levels.
- That is, we retain the hypothesis that student status does
not modify the effect of income on balance.


## Statistical Interaction More Generally

- Interaction terms are generally defined as products of two other model terms:
    - $f(X) = \beta_0 + \beta_1 X_1 + \beta_2X_2 + \beta_3 X_1X_2$
- In general we allow different lines for different values of $X_2$.
    - $X_2 = 0$ model: intercept $\beta_0$ and slope $\beta_1$
    - $X_2 = x_2$ model: intercept $\beta_0+\beta_2 x_2$ and slope $\beta_1+\beta_3x_2$
    - Interpret $\beta_3$ as the difference between slopes for a one-unit change in $X_2$.
    - If $\beta_3=0$ then $X_2$ does not modify effect of $X_1$ on $Y$ and _vice versa_.
    - In practice, test the hypothesis $H_0:\beta_3=0$.


## Interaction Between TV and Radio Advertising

\scriptsize

```{r}
afit <- lm(sales ~ TV*radio,data=advert)
summary(afit)$coefficients
```

\normalsize

* There is statistical interaction between TV and radio ads.
    + TV modifies the effect of radio, or radio modifies the
    effect of TV

# Categorical Variables as Predictors

## Dummy Variables for Categorical Predictors

- We have seen dummy variables before.
    - A binary variable for student status,
    coded 0 or 1 to represent 
    the two levels of a dichotomous variable.
- When the categorical variable has more than two values,
or levels, we need more than one binary "dummy" variable.
    - Example: The `Ethnicity` variable from the 
    credit data.

## 

\scriptsize

```{r, fig.width=6,fig.height=2.5}
ggplot(credit,aes(x=Income,y=Balance,color=Ethnicity)) +
  geom_point() + geom_smooth(method="lm",se=FALSE)
```

## Regression Model for Balance

\scriptsize

```{r}
cfit <- lm(Balance ~ Income*Ethnicity,data=credit)
summary(cfit)$coefficients
```





## Model Details

- Model requires too much notation to write out in detail;
will explain in the context of this example.

\scriptsize

```{r}
coefficients(cfit)
```

\normalsize

- The model uses African American as a "baseline".
    - The intercept=175 and slope `Income`=7.46 
    terms are the fitted model for mean `Balance` in African
    Americans.
- The model for mean `Balance` in another ethnic group is 
the baseline plus ethnic-group-specific intercept and slope
    - E.G., for the Asians, add 57.3 to the African American
    intercept and $-1.13$ to the African American slope
    
## Dummy Variables for Ethnic Group

\small

- Create a binary variable for each non-baseline ethnic group
that takes value 1 if the person is from that ethnic group and 0
otherwise.

\scriptsize

```{r, echo=FALSE}
dvars <- model.matrix(cfit)[,2:4]
head(data.frame(Ethnicity=credit$Ethnicity,dvars))
```


## Model with separate lines for each continent

\scriptsize

```{r, echo=FALSE}
dvars <- model.matrix(cfit)[,2:6]
head(data.frame(Ethnicity=credit$Ethnicity,dvars))
```


## Multiple-Partial F-test for Interaction

\scriptsize

```{r}
cfitReduced <- lm(Balance ~ Income + Ethnicity,data=credit)
anova(cfitReduced,cfit)
```

\normalsize

* There is little evidence that ethnicity modifies
the effect of income on balance.


# Model Diagnostics based on Residuals

## Model Diagnostics based on Residuals

* Residuals are the primary tool for 
    + checking model assumptions (correct linear
    model, constant error SD,  and normal errors) and
    + identifying unusual observations.
* Residuals may also be useful for detecting correlation 
in the errors, but this is a more specialized topic not discussed.

## Residuals versus fitted values


-  Plot residuals _vs_ fitted values
to assess adequacy of the linear model and constant error
SD.
    - A plot of $\epsilon_i$ _vs_ $f(x_i)$ would show 
    no pattern, because the $\epsilon_i$'s are random noise.
    - If the linear model is adequate, we should not see any 
    trends or patterns in the residuals _vs_ fitted values
    $\hat{y}_i = \hat{f}(x_i)$.
    - Also, if the error SD is constant, the variation in 
    residuals _vs_ $\hat{y}_i$  should
    look roughly equal.
- We may also see outliers in the regression sense.

## Residuals versus Fitted Values - Advertising

* Use the `residual()` and `fitted()` extractor functions.

\scriptsize

```{r, fig.width=4, fig.height=2.5}
adAug <- data.frame(advert,fitted=fitted(afit),residuals=residuals(afit))
ggplot(adAug,aes(x=fitted,y=residuals)) +
  geom_point() + geom_smooth()
```



## Residual vs fitted -- comments

\small

- Horizontal line at zero is outside the error bands around smoother line. 
    + Suggests we have missed a non-linear trend.
- Spread of residuals fairly constant over range of fitted values,
so constand SD assumptions appears reasonable.


## Q-Q Plots

\small

- A quantile-quantile (Q-Q) plot is a plot of  the quantiles of one distribution 
to another.
    - If the two distributions have the similar shape, the points should fall
roughly on a straight line.
- Our interest is in comparing the quantiles of the distribution of 
residuals to the quantiles of the distribution they should have under normal errors.
    - One can argue that the residuals don't have 
    the same distribution (those closer to the centre of the
    plot are slightly more variable).
- However, Studentized residuals do -- they have a $t$ distribution with 
$n-k-2$ degrees of freedom.
    - The Studentized residual for the $i$th case
is  
$$ \frac{y_i - \hat{y}_i}{\hat{\sigma}_{-i}\sqrt{1-h_{ii}}}.$$
    where $h_{ii}$ is the leverage or hat value and 
    $\hat{\sigma}_{-i}$ is the estimate of the error SD **without** case $i$
    
## Q-Q plot of Studentized residuals

\small

- The `qqPlot()` function from the `car` package plots
Studentized residuals against quantiles of the appropriate
$t$ distribution.
    - Also adds error bands: If all points fit within bands, 
    it is plausible that the sample is from the $t$.

\tiny

```{r, fig.width=4, fig.height=2.5, warning=FALSE,message=FALSE}
library(car) # Use install.packages("car") to install
qqPlot(afit)
```



## Identifying unusual observations

- Studentized residuals can identify outliers
    - Rule-of-thumb: Residuals beyond $\pm 2$ are moderate outliers and beyond $\pm 3$ are serious outliers.
- Leverage ($h_i$) is a measure of how atypical an observation's
$X$ values are.
    - Rule-of-thumb: $h_i > 2(p+1)/n$ is somewhat high leverage
    and $h_i > 3(p+1)/n$ is very high leverage.
- Cook's distance measures the influence of an observation; i.e., how much the estimated regression coefficients change when the observation is removed.
    - Rule-of-thumb: Cook's distance $> 0.5$ is moderately
    influential, and $> 1$ is highly influential


## Identify Unusual Observations in Advertising Data

\small

- Augment the dataset and `View()` to identify
cases that are
unusual according to our rules-of-thumb.
    - Several "moderate" residuals, and two severe
    residuals of $-4.5$ and $-7.9$.
    - For leverage, $p=2$, $n=200$, $2(p+1)/n = 0.03$,
    $3(p+1)/n=0.045$: 27 cases with moderate leverage, 12 with 
    very high leverage!
    - One moderately influential case.

```{r}
adAug <- data.frame(advert,studRes = rstudent(afit),
                    hats = hatvalues(afit), 
                    cooks = cooks.distance(afit))
# Now View(adAug)
```

## Correlated Predicters, or Collinearity

- The distribution of $X$'s can affect stability of the 
least squares estimates.
    - For simple linear regression one can show that: 
    $$
    SE(\hat{\beta}_1) = \frac{\hat{\sigma}}{S_X \sqrt{n-1}}
    $$
    where $S_X$ is the SD of the $X$'s in the dataset.
    - Implies that the larger the $S_X$ the smaller the SE
    (i.e., the more stable the fit).
- In general can think of the positioning of $X$'s as a "foundation" that
supports the least squares surface -- the broader the base, 
the more stable the estimates.
- Collinearity, or correlation between predictors, yields an unstable 
foundation and hence unstable estimates.

## More on SEs

\small

- With two explanatory variables, $X_1$ and $X_2$, 
can show that
$$ 
SE(\hat{\beta}_1) = \frac{\hat{\sigma}}{S_{X_1} \sqrt{n-1}\sqrt{1-r^2_{12}}}
$$
and
$$ 
SE(\hat{\beta}_2) = \frac{\hat{\sigma}}{S_{X_2} \sqrt{n-1}\sqrt{1-r^2_{12}}}
$$
where $r_{12}$ is the correlation between $X_1$ and $X_2$. 
- In addition to $S_{X_1}$ and $S_{X_2}$ we must consider 
the correlation between $X_1$ and $X_2$.
- The larger the squared correlation, the larger the SEs

## Variance Inflation Factors (VIFs)

\small

- In a multiple regression with $X_1,\ldots,X_p$,  
$$ 
SE(\hat{\beta}_j) = \frac{\hat{\sigma}}{S_{X_j} \sqrt{n-1}\sqrt{1-R^2_j}}
$$
where $R^2_j$ is the $R^2$ from the regression of $X_j$ on $X_{(-j)}$.
- The term $1/\sqrt{1-R^2_j}$,
is the factor by which the SE of $\hat\beta_j$
is inflated over the SE from a simple linear regression
by correlation between $X_j$ and the other $X$'s.
- The variance inflation factor for $X_j$, $VIF_j$, is
defined to be $1/(1-R^2_j)$; i.e., the SE of $\hat\beta_j$
is inflated by $\sqrt{VIF_j}$.
- High VIFs indicate instability.
    - One rule of thumb is that a $VIF_j> 10$ is cause for concern.
    
## VIFs and Other Diagnostics in the `car` Package

\small

- If you haven't already done so, install the 
R package `car`.

\scriptsize

```{r}
library(car)
vif(afit)
```

\normalsize

* The VIFs suggest the interaction is quite well predicted
by TV and radio, but the VIF is less than our threshold 
so we are not concerned.

## Collinearity with Polynomial Terms

\scriptsize

```{r}
uu <- url("http://www-bcf.usc.edu/~gareth/ISL/Income1.csv")
income <- read.csv(uu,row.names=1)
ifit<- lm(Income ~ Education + I(Education^2) + I(Education^3), data=income)
vif(ifit)
```

## Remedies for collinearity

- When the collinearity arrises from explanatory variables that 
are products of other variables, centering can help.

\scriptsize

```{r}
centre <- function(x) { x - mean(x)}
income <- data.frame(income, cEducation = centre(income$Education))
ifit<- lm(Income ~ cEducation + I(cEducation^2) + I(cEducation^3), data=income)
vif(ifit)
```


## Collinearity: If centering doesn't help

- May need to exclude a variable.
    - Sounds drastic, but high $R^2_j$ indicates $X_j$ is very well predicted by $X_{(-j)}$, so nothing really lost.
- Which variable to exclude? 
- First use common sense: 
    - If one variable is 
    a surrogate for another, drop the surrogate.
    - For example, if we are modeling house prices
    with (i) size of the house in square feet 
    and (ii) the number of bedrooms, we may think
    bedrooms is just a surrogate for size.
- If no obvious candidate to drop, use model selection.

## Fit of the Advertising Data Revisited

\scriptsize

```{r}
library(dplyr)
advert <- mutate(advert,cTV = TV - mean(TV),
         cRadio = radio - mean(radio))
afitC <- lm(sales ~ cTV*cRadio+I(cTV^2) +I(cTV^3)+I(cTV^4),
            data=advert)
summary(afitC)$coefficients
vif(afitC)
```


##

\scriptsize

```{r, fig.width=4, fig.height=2.5}
mutate(advert,fitted=fitted(afitC),residuals=residuals(afitC)) %>%
  ggplot(aes(x=fitted,y=residuals)) + geom_point() + geom_smooth()
```

