---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 7, Part 4: Generalized Additive Models'
author: "Brad McNeney"
date: '2017-10-31'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Generalized Additive Models (GAMs)

* We now consider extending the linear model 
when we have $p$ explanatory variables,
$X = (X_1,\ldots,X_p)$. 
* In linear regression, the function $f(X)$ is of the form
$$f(X) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \ldots + \beta_p X_p$$
* In a GAM we use (up to) $p$ smooth functions
$$f(X) = \beta_0 + f_1(X_1) + f_2(X_2) + \ldots + f_p(X_p)$$
* The component functions $f_j(\cdot)$ can be any
of the smoothers discussed in Sections 7.1-7.6
(e.g., polynomial, spline or local regression; smoothing spline)

## Example: Wage Data (Again)

* Fit a model for `wage` of the form
$$wage = \beta_0 + f_1(year) + f_2(age) + f_3(education) + \epsilon$$
* Recall that `education` is categorical, so 
$f_3$ is an expansion into dummy variables.

\scriptsize

```{r}
library(ISLR)
data(Wage)
table(Wage$education)
```

\normalsize

* We can use natural cubic splines with 4 df in `year` and `age`.

\scriptsize

```{r}
library(splines)
gfit <- lm(wage ~ ns(year,4) + ns(age,4) + education,data=Wage)
```

## 

* To plot we can use a plotting function from the 
`gam` package.

\scriptsize

```{r}
library(gam)
par(mfrow=c(1,3))
plot.gam(gfit, se=TRUE)
```

## Model Selection

* We could do CV-based model selection on the df for the two 
splines, but this would now be a search over a 2-d 
grid of df's.
* However, we notice that a linear fit in `year` looks
plausible, and we can use an ANOVA F-test to test the null
hypothesis of linearity.
    + Importantly, the model that is linear in `year` 
    is a sub-model of the natural cubic spline model
    (spline with 1 df is linear)

\scriptsize

```{r}
gfit2 <- lm(wage ~ year+ns(age,4)+education,Wage)
anova(gfit2,gfit)
```

\normalsize

* We retain the hypothesis that $f$ is linear in `year`.

## GAM with Smoothing Splines

* Smoothing splines shrinkage estimators
and so are not fit simply by least squares.
* Use the `gam` package.
    + Written by Hastie and Tibshirani (also authors
    of a book on the subject)
    
\scriptsize

```{r}
library(gam)
gfit3 <- gam(wage ~  s(year,4) + s(age,4) + education,data=Wage)
```

##

\scriptsize

```{r}
par(mfrow=c(1,3))
plot(gfit3, se=TRUE,ylim=c(-40,40))
```

## GAM Intepretation

* Each smooth is the estimated effect of changing
one variable holding the others fixed.
* For example, holding age and education fixed,
wage increases slightly, and approximately linearly
by year.
* Holding year and education fixed,
wage increases until about 40, then is levels out,
and the drops after 60.
* Holding year and age fixed, 
wage increases with education level.


## Model Reduction

* A smoothing spline with 2df in `year` is linear, so we can 
use the F-test to compare models.

\scriptsize

```{r}
gfit4 <- gam(wage ~ + year + s(age,4) + education,data=Wage)
anova(gfit4,gfit3)
```

\normalsize

* We retain the hypothesis of linear in `year`.

##

\scriptsize

```{r}
par(mfrow=c(1,3))
plot(gfit4,se=TRUE,ylim=c(-40,40))
```

## Model Summary

\tiny

```{r}
summary(gfit4)
```

## GAM Predictions

\scriptsize

```{r}
newdat <- expand.grid(
  year=2003:2009,
  age=c(20,30,40,50,60,70,80),
  education=levels(Wage$education))
preds <- predict(gfit4,newdata=newdat)
preds[,,5] # Advanced degree
```

## GAM with Local Regression

\scriptsize

```{r}
gfit5 <- gam(wage ~ year + lo(age,span=0.5)+education,data=Wage)
par(mfrow=c(1,3))
plot(gfit5,se=TRUE)
```

## GAM with Multiple Local Regression ("Interaction")

\scriptsize

```{r}
gfit6 <- gam(wage ~ lo(year,age,span=0.5)+education,data=Wage)
par(mfrow=c(1,2))
library(akima)
plot(gfit6,se=TRUE)
```


## Advantages and Limitations of GAMs

* Advantages
    + Allows non-linear relationships that simple
    linear regression might miss, or might take a lot 
    of work to discover (think age effect). 
    + Can interpret components of the GAM
    (holding other variables fixed)
    + Smoothness of the component functions can 
    be controlled by their df.
* Disadvantages
    + Restricted to additive models, though 
    can fit interactions with local regression.

## GAMs for Classification

* The "G" in GAM also stands for a generalization
beyond gaussian linear models.
* Recall the logistic regression 
model formulation for modelling 
$p(X) = P(Y=1|X)$:
$$ \log \left( \frac{p(X)}{1-p(X)} \right) = 
\beta_0 + X_1 \beta_1 + \ldots + 
X_p \beta_p. $$
* Generalize to 
$$ \log \left( \frac{p(X)}{1-p(X)} \right) = 
\beta_0 + f_1(X_1) + \ldots + f_p(X_p). $$

## Example Logistic GAM

\scriptsize

```{r}
gfit7 <- gam(I(wage>250) ~ year + s(age,4) + education,
             data=Wage,family=binomial)
par(mfrow=c(1,3))
plot(gfit7,se=TRUE,ylim=c(-6,6))
```

## Removing `< HS Grad`

* There are no high-income earners with `< HS Grad` education,
so our esimate of the education effect in this category 
is essentially $-\infty$.
    + Remove this category and re-fit
    
\scriptsize

```{r}
gfit7.s <- gam(I(wage>250) ~ year + s(age,4) + education,
             data=Wage,family=binomial,subset=(education!="1. < HS Grad"))
```

##

\scriptsize

```{r}
par(mfrow=c(1,3))
plot(gfit7.s,se=TRUE,ylim=c(-6,6))
```

## Remove Year

\scriptsize

```{r}
gfit8.s <- gam(I(wage>250) ~  s(age,4) + education,
             data=Wage,family=binomial,subset=(education!="1. < HS Grad"))
anova(gfit8.s,gfit7.s)
```

## Alternative Implementation of GAMs

* `mgcv` is another well-developed R package that
fits GAMS.
* The focus in `mgcv` is on penalized regression splines.
    + Penalty term may be selected by CV or other estimate of 
    test set error.
* Allows interactions through "thin plate" regression splines
