---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 7, Part 2: Basis Functions and Regression Splines'
author: "Brad McNeney"
date: '2017-10-24'
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Basis Functions

* The polynomial of degree 4 that we fit to the wage data
is a function $f(X)$ of the form
$$f(X) = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 X^4$$
    + Think of this as a linear combination of four functions,
    $b_1(x) = x$, \ldots, $b_4(x) = x^4$.
    + The four functions are called *basis functions*
* In this chapter we fit functions $f(X)$ of the form
$$f(X) = \beta_0 + \beta_1 b_1(X) + \ldots, + \beta_k b_K(X)$$
for some choice of basis functions.

## Visualizing Basis Functions

\scriptsize

```{r}
library(ISLR); data(Wage)
age <- sort(unique(Wage$age)); age <- scale(age)
Xmat <- data.frame(age=age,b1=age,b2=age^2,b3=age^3,b4=age^4)
library(tidyr); library(ggplot2)
Xlong <- gather(Xmat,basisFun,b,-age) 
ggplot(Xlong,aes(x=age,y=b,color=basisFun)) + geom_line()
```

## Basis Functions from `poly()`

\scriptsize

```{r}
age <- as.numeric(age) #scaling made it unsuitable as input for poly()
Xmat <- data.frame(age=age,poly(age,4))
Xlong <- gather(Xmat,basisFun,b,-age) 
ggplot(Xlong,aes(x=age,y=b,color=basisFun)) + geom_line()
```

## Step Functions as Basis Functions

* The step functions we used to model wages can also 
be considered basis functions.
* They are defined to be "piece-wise constant"; that is,
constant on the separate pieces of the x-axis defined
by the age category cut-offs.
    + We will soon refer to the "interior" cut-offs as knots.


\scriptsize

```{r}
age <- sort(unique(Wage$age)) # return to original age
# Use min, max and quartiles of age distribution.
agebreaks <- quantile(Wage$age) 
# Nuisance: need to widen left-most cutpoint to include min
agebreaks[1]<-agebreaks[1]-1
# Use model.matrix() to set up the dummy vars. Exclude Intercept
Xmat <- model.matrix( ~ cut(age,agebreaks), data=data.frame(age=age))[,-1] 
Xmat <- data.frame(age,Xmat)
Xlong <- gather(Xmat,basisFun,b,-age) 
```

##

\scriptsize

```{r}
ggplot(Xlong,aes(x=age,y=b,color=basisFun)) + geom_line()
```

## Piecewise Polynomials

* Instead of constants between the cut-points, we can use
polynomials.
    + Start with cubic polynomials.
* Then impose constraints on the polynomials
that guarantee a smooth function whenever we take a 
linear combination (more on this later).
    + With these constraints, it turns out that
    $K$ interior knots leads to $K+3$ basis functions.
    + For later: $K+4$ coefficients, or degrees of freedom (df).

\scriptsize


```{r}
library(splines)
knots <- agebreaks[2:4] #interior cut-points
Xmat <- data.frame(age=age,bs(age,knots=knots))
Xlong <- gather(Xmat,basisFun,b,-age) 
```

##

\scriptsize

```{r}
ggplot(Xlong,aes(x=age,y=b,color=basisFun)) + geom_line() +
  geom_vline(xintercept=knots)
```

\normalsize

* Note: In the end, the basis functions are not
restricted to the intervals between/outside the knots

## Constraints Make Smooth Functions

* If we take any linear combination of 
the basis functions we always get a smooth curve.
    + No jumps, **and** no obvious "kinks".
  
\scriptsize

```{r}
Xmat <- bs(age,knots=knots)
coefs <- rnorm(ncol(Xmat)) # arbitrary coefficients -- try your own
dat <- data.frame(age=age,f =Xmat%*%coefs)
```

##

\scriptsize

```{r}
ggplot(dat,aes(x=age,y=f)) + geom_line() +
  geom_vline(xintercept=knots)
```

## Fitting a Regression Spline to the Wage Data

* With the basis functions in hand, we use ordinary
least squares to fit the regression spline.

\scriptsize

```{r}
fit <- lm(wage ~ bs(age,knots=knots),data=Wage)
summary(fit)$coef
plotfit <- function(fit,dat,newdat){
  preds <- data.frame(newdat,
          predict(fit,newdata=newdat,interval="confidence"))
  ggplot(dat,aes(x=age,y=wage)) + geom_point(alpha=0.1) +
    geom_ribbon(aes(x=age,y=fit,ymin=lwr,ymax=upr),
                data=preds,fill="blue",alpha=.2) + 
    geom_line(aes(y=fit),data=preds,color="blue")
}
```

## Plotting Fitted Curve

\scriptsize

```{r}
newdat <- data.frame(age=age)
plotfit(fit,Wage,newdat)
```

## Constraints Outside the Knots

* Estimated splines curves have high variance where
data are scarce.
* Illustrate by sampling from the wage data.

\scriptsize

```{r,fig.height=2.5}
Wage.sub <- rbind(Wage[sample(1:nrow(Wage),50),c("wage","age")],
                  c(100,75))
fit <- lm(wage ~ bs(age,knots=knots),data=Wage.sub)
plotfit(fit,Wage.sub,newdat)
```

## Natural Splines

* Natural splines constrain the function to be linear 
outside the knots, and these constraint has the effect
of reducing variance.
    + Note on df: two extra constraints means two fewer df;
    i.e., $K+1$.

\scriptsize

```{r,fig.height=2.5}
fit <- lm(wage ~ ns(age,knots=knots),data=Wage.sub)
plotfit(fit,Wage.sub,newdat)
```

## Knots or Degrees of Freedom

* We mentioned that $K$ interior knots leads to 
$K+3$ df for the cubic spline and $K+1$ for the
natural spline. 
* Can specify df and let `bs()` or `ns()` choose
the knots at appropriate quantiles of `age`.
* Why? A single smoothing parameter lends itself to 
cross-validation to select it.

##

\scriptsize

```{r}
cv <- function(df,k,seed) {
  # Setup
  set.seed(seed)
  folds <- sample(1:k,size=nrow(Wage),replace=TRUE)
  validErr <- rep(NA,k)
  # Loop over folds: train, test, MSE
  for(i in 1:k) {
    testWage <- Wage[folds==i,]
    trainWage <- Wage[folds!=i,]
    fit <- lm(wage ~ ns(age,df=df),data=trainWage)
    testPreds <- predict(fit,testWage)
    validErr[i] <- mean((testPreds - testWage$wage)^2)
  }
  mean(validErr)
}
k<-10; nDf <- 10; seed <- 1; cvErrs <- rep(NA,nDf)
for(df in 1:nDf){ # loop over df
  cvErrs[df] <- cv(df,k,seed)
}
cvErrs
```

##

\scriptsize

```{r,fig,height=5}
plot(1:nDf,cvErrs,type="b",ylab="MSE",xlab="DF for natural spline")
```

## Conclusion for Wage Data

* About three df looks adequate.

\scriptsize

```{r}
fit <- lm(wage ~ ns(age,df=3),data=Wage)
plotfit(fit,Wage,newdat)
```


