---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 4, Part 1: Introduction and Logistic Regression'
author: "Brad McNeney"
date: '2018-09-24'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

# Introduction

## Classification Problems

* Instead of a quantitative response, we have a 
categorical (qualitative) response, such as yes/no.

\scriptsize

```{r}
library(ISLR)
data(Default)
head(Default)
```

## Classification Problems, cont.

* Predicting a categorical response is 
called classifying.
    + We are assigning the observation to a category
    or class.
* Classification methods may be based on 
modelling the probability of class membership.
    + Assign to class with highest probability
    + Modelling class probabilities can be 
    cast as a regression.
* Most popular classifiers are 
logistic regression, linear discriminant analysis
and K-nearest neighbors.

## Overview of Classification

* Will use a set of training observations, $\{ (x_1,y_1),\ldots,(x_n,y_n) \}$ to build the 
classifier.
* Use the `Default` data to illustrate.
* The response is `default` on credit card 
payment (Yes/No), to 
be predicted by credit card `balance`, annual `income`
and `student` status (Yes/No).

\scriptsize

```{r}
summary(Default)
```


## Default data

\scriptsize

```{r}
dtab <- xtabs(~ default + student,data=Default)
dtab
prop.table(dtab,margin=2)
```

##

\scriptsize

* Overplotting is a problem with a data set this large.

```{r,fig.height=3,fig.width=5}
library(ggplot2)
ggplot(Default,aes(x=balance,y=income,color=default)) + geom_point(alpha=0.1)
```

## Why Not Linear Regression?

* Numerical codings of categorical 
variables have no real meaning. 
    + Recall `origin` variable in `Auto` data.
* What does it mean for a one unit
increase in $X_i$ to be associated with a $\beta_i$
increase in a categorical response?
* Binary response (success/failure) may be the exception.
    + Say we code success=1, failure=0.
    + Can show that a linear regression 
    predicts the probability of success given $X$.
    + But probabilities not constrained to be 
    between 0 and 1.

# Logistic Regression

## Notation and Use as Classifier

* Model $Pr({\rm success} | X)$ as a function $p(X)$ 
of $X$. 
    + E.G., for the `Default` data, model
    $Pr({\rm default=yes}|{\rm balance}) = p({\rm balance})$.
* Predict response for new $x_0$ to be success if $p(x_0)>c$
for some constant $c$, such as 1/2.

## The Logistic Model.

* Model is linear in the log-odds (logit) of success:
$$ \log \left( \frac{p(X)}{1-p(X)} \right) = 
X \beta = \beta_0 + X_1 \beta_1 + \ldots + 
X_p \beta_p. $$
    + A one unit increase in $X_j$ holding
    all others fixed is associated with a 
    $\beta_j$ change in the log-odds.
* Can show that the logit model implies $p(\cdot)$ is the logistic
function of $X\beta$:
$$ p(X) = \frac{e^{X\beta}}{1+e^{X\beta}}. $$

## The Logistic Function

\scriptsize

```{r}
seqLen <- 100
x <- seq(from=-5,to=5,length=seqLen)
dd <- data.frame(x=x,logist = exp(x)/(1+exp(x)))
ggplot(dd,aes(x=x,y=logist)) + geom_line()
```

## Estimating $\beta$ by Maximum Likelihood

* Choose $\hat{\beta}$ to maximize the likelihood.
* The likelihood is the probability of the 
observed data, viewed as a function of $\beta$.
* We assume independent observations, which means
the probability of the data is the product of 
the probabilities of each observation.
* For the $i^{th}$, the probability of a success is $p(x_i)$
and the probability of a failure is $1-p(x_i)$.  
* Thus 
$$ L(\beta) = \prod_{i:y_i=1} p(x_i) \prod_{i:y_i=0} (1-p(x_i)) = \prod_i p(x_i)^y_i (1-p(x_i))^{1-y_i}$$

## Maximizing the Log-Likelihood

* The maximizer of the likelihood is the same as the 
maximizer of the log-likelihood

$$ l(\beta) = \log L(\beta) = \sum_i y_i \log p(x_i) 
+ (1-y_i) \log (1-p(x_i)). $$

* In many problems, maximizing the log-likeihood
is an easier optimization problem.

## Fitting a Logistic Regression in R

* Use the `glm()` function (generalized
linear models, with logistic as a special case).

\scriptsize

```{r}
dfit <- glm(default ~ balance + income + student,
            data=Default, family=binomial())
round(summary(dfit)$coefficients,4)
```

\normalsize

* Income does not predict default; no evidence
of interaction between students status and balance (not shown)


## Software notes

* `glm()` interface is very similar to `lm()`.
* New argument `family` specifies the type of 
GLM.
    + `binomial()` is for binary outcomes, or 
    outcomes that are sums of binary variables.
    

## Confounding

\scriptsize

```{r}
dfitS <- glm(default ~ student,data=Default,family=binomial())
dfitSB <- glm(default ~ student+balance,data=Default,family=binomial())
round(coefficients(dfitS),5); round(coefficients(dfitSB),5)
```

\normalsize

* Including balance reverses the effect of student:
    + Without balance, students look more likely
    to default (why?).
    + Adjusting for balance, students are less likely
    to default; i.e., given a student and a non-student
    with the same balance, the student is less likely
    to default.

## We Do Not Test for Confounding

* Find the percentage change in the coefficient with
and without balance:
$$ \frac{0.405 - (-0.715)}{|-0.715|} \times 100\%
= 156.6 $$
* If the change is more than some threshold (e.g., 
10%) we say balance confounds the association
between default and student.
* By contrast, we **do** test for interaction.


## Cause of Confounding

* For a variable like balance to confound the 
association between default and student, 
it must be associated with both.
    + Higher balance, higher default rate.
    + Higher balance, more likely student.
    + Looks like students have higher default rate.

##

\scriptsize

```{r}
ggplot(Default,aes(x=balance,fill=default)) + geom_histogram()
```

## 

\scriptsize

```{r}
ggplot(Default,aes(x=balance,fill=student)) + geom_histogram()
```


## Predictions of $p(X)$.

* Plug in the values of a new $x_0$ into the
fitted equation to get $\hat{p}(x_0)$.

\scriptsize

```{r}
bal <- seq(from=min(Default$balance),to=max(Default$balance),len=seqLen)
newdat <- data.frame(balance = c(bal,bal),
                     student = factor(c(rep("Yes",seqLen),
                                        rep("No",seqLen))))
pred <- data.frame(newdat,
                   pred = predict(dfitSB,newdata=newdat,type="response"))
head(pred)
```

##

\scriptsize

```{r}
ggplot(pred,aes(x=balance,y=pred,color=student)) + geom_line()
```

## Logistic Regression for $>2$ Response Categories

* Instead of a logistic model we fit a polytomous
or multinomial logistic regression.
* Functions such as `multinom()` in the 
`nnet` package can fit.
    + We won't go into details.
* Text suggests such models are less popular 
than discriminant analysis, to be discussed in 
the next set of notes.
