---
title: 'Statistics 452: Statistical Learning and Prediction'
subtitle: 'Chapter 4, Part 2: Linear Discriminant Analysis Classifier'
author: "Brad McNeney"
date: '2018-09-25'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

# Linear Discriminant Analysis (LDA) Classifier

## Logistic Regression vs LDA Classifieer

* Logistic regression models $Pr(Y=1|X=x) = p(x)$.
    + More generally, polytomous regression models $Pr(Y=k|X=x)$.
* LDA models $Pr(X=x|Y=k)$, which relates to 
$Pr(Y=k|X=x)$ _via_ Bayes' rule:
$$Pr(Y=k|X=x) = \frac{Pr(X=x|Y=k)Pr(Y=k)}{\sum_j Pr(X=x|Y=j)Pr(Y=j)}.$$

## Notation and Terminology

* Notation: $f_k(X) = Pr(X=x|Y=k)$,
$\pi_k = Pr(Y=k)$, $p_k(x) = Pr(Y=k|X=x)$,  so that 
$$p_k(x) = \frac{f_k(x)\pi_k}{\sum_j f_j(x)\pi_j}.$$
* Terminology: 
    + $\pi_k$ is the _prior_ probability of class $k$
    + $p_k(x)$ is the _posterior_ probability of
    class $k$ given data $x$.

## Bayes Classifier

* We saw in Chapter 2 that the classifier with 
the lowest error rate is the one that 
chooses the class with the highest posterior probability.
* If we have good estimates, $\hat{p}_k(x)$, 
we can choose the class with the highest **estimated**
posterior probability.
* Estimate $p_k(x)$ by estimating
    + $\pi_k$: If training sample is a population 
    random sample, use the sample proportions.
    + $f_k(x)$: LDA model is parametric, so 
    we estimate $f_k$ by estimating its parameters. 
    
## LDA Classifier: Model for $f_k(x)$ when $p=1$

* Assume $f_k(x)$ is normal with mean 
$\mu_k$ and variance $\sigma_k^2$. 
    + Then estimating $f_k$ amounts to estimating 
    $\mu_k$ and $\sigma_k^2$.
    + Further assume $\sigma_k$ is the same 
    for all $k$ and let $\sigma$ denote this common value.
* For a particular $x$,
the class with highest posterior probability is 
the one with highest value of
$$\delta_k(x) = x \frac{\mu_k}{\sigma^2}  - 
\frac{\mu_k^2}{2\sigma^2} + \log \pi_k.$$

## Example

* Suppose two classes, $\pi_1=0.1$, $\pi_2=0.9$,
$\mu_1=1$, $\mu_2=-1$, $\sigma=1$.

\scriptsize

```{r}
mu1<-1; mu2<-(-1); sigma<-1; pi1<-0.1;pi2<-0.9
x <-seq(from=-4,to=4,length=100)
f1 <- dnorm(x,mean=mu1,sd=sigma)
f2 <- dnorm(x,mean=mu2,sd=sigma)
delta <- function(x,mu,sigma,pi) {
  x*mu/sigma^2 - mu^2/(2*sigma^2) + log(pi)
}
delta1 <- delta(x,mu1,sigma,pi1)
delta2 <- delta(x,mu2,sigma,pi2)
dd <- data.frame(x=x,f1=f1,f2=f2,delta1=delta1,delta2=delta2)
```

##

\scriptsize

```{r}
library(tidyverse)
ggplot(dd,aes(x=x))+
  geom_line(aes(y=f1*pi1),color="cyan")+
  geom_line(aes(y=f2*pi2),color="red") + 
  labs(y=expression(f[k](x)*pi[k]))
```

## 

\scriptsize

```{r}
ggplot(dd,aes(x=x))+
  geom_line(aes(y=delta1),color="cyan")+
  geom_line(aes(y=delta2),color="red") + 
  labs(y=expression(delta[k](x)))
```

## Decision Boundary

* The point where the $\delta_k$ 
lines cross is the decision 
boundary.
* Can show that the boundary is 
$$\frac{\mu_1 + \mu_2}{2} + \frac{\sigma^2\log(\pi_2/\pi_1)}{\mu_1-\mu_2}.$$
* In this example the boundary is 
$(1 + (-1))/2 + \log(.9/.1)/2 \approx 1.1$

\scriptsize

```{r}
(mu1 + mu2)/2 + sigma^2*log(pi2/pi1)/(mu1-mu2)
```

## Estimated Distributions

* Replace means and variance in $\delta_k(x)$'s  with estimates (equation 4.15 of text) to get the _discriminant functions_
$\hat{\delta}_k$.
* Illustrate with `Default` data.

\scriptsize

```{r}
library(ISLR)
data(Default)
```

##

\scriptsize

```{r}
ggplot(Default,aes(x=balance,color=default)) +
  geom_density(aes(y=..count..))
```

## Software Note

* Specifying `color=default` groups the data 
by `default`.
* The density geom plots the estimated densities
$\hat{f}_k(x)$ by default, but also computes the variable
`..count..`, which for group $k$ is
$\hat{f}_k(x) n_k$.
* We are interested in $\hat{f}_k(x) \hat{\pi}
= \hat{f}_k(x) n_k/n$,
so  `count` is proportional to what we want to plot.

## Estimated Densities and Discriminant Functions

\scriptsize

```{r}
n <- nrow(Default); K <- 2
pi1 <- with(Default,mean(default=="Yes")); pi2 <- 1-pi1
defBalance <- with(Default,balance[default=="Yes"])
nodefBalance <- with(Default,balance[default=="No"])
mu1 <- mean(defBalance)
mu2 <- mean(nodefBalance)
sigma <- sqrt((sum((defBalance-mu1)^2) + sum((nodefBalance-mu2)^2))/(n-K))
with(Default,range(balance))
x <- seq(from=0,to=3000,length=100)
f1 <- dnorm(x,mean=mu1,sd=sigma)
f2 <- dnorm(x,mean=mu2,sd=sigma)
delta1 <- delta(x,mu1,sigma,pi1)
delta2 <- delta(x,mu2,sigma,pi2)
dd <- data.frame(x=x,f1=f1,f2=f2,delta1=delta1,delta2=delta2)
```

##

\scriptsize

```{r}
ggplot(dd,aes(x=x))+ geom_line(aes(y=f1*pi1),color="cyan")+
  geom_line(aes(y=f2*pi2),color="red") + 
  labs(y=expression(hat(f)[k](x)*hat(pi)[k]))
min(x[f1*pi1>f2*pi2])
```


##

\scriptsize

```{r}
ggplot(dd,aes(x=x))+
  geom_line(aes(y=delta1),color="cyan")+
  geom_line(aes(y=delta2),color="red") + 
  labs(y=expression(hat(delta)[k](x)))
```

## Decision Boundary for Default Data

\scriptsize

```{r}
(mu1 + mu2)/2 + sigma^2*log(pi2/pi1)/(mu1-mu2)
```

## LDA Classifier in R

\scriptsize

```{r}
library(MASS)
ll <- lda(default ~ balance, data=Default)
preds <- predict(ll)
head(preds$posterior) 
head(preds$class)
```

## LDA Classifier for $p>1$

* Same basic idea of classifying based 
on highest posterior probabilities.
* Generalize the model for $f_k(X)$
from a Gaussian distibution to 
a _multivariate_ Gaussian distribution.
* The mean in class $k$ is now a vector
$\mu_k$ and the variance is a matrix $\Sigma$.
* Details on the multivariate density and 
the resulting $\delta_k(x)$ are given on page 143
of the text.

## LDA Classifier Using all Default Data

\scriptsize

```{r}
ll <- lda(default ~  student + balance + income, data=Default)
preds <- predict(ll)
head(preds$posterior)
```


## Predictions _vs_ True Default Status

* Tabulating predictions and true default status
gives the following "confusion matrix":

\scriptsize

```{r}
Default <- data.frame(Default,predicted = preds$class)
xtabs( ~ predicted + default, data=Default)
```

\normalsize

* (Slight difference with results in 
text.)
* 9724 correctly predicted, 276 incorrectly 
predicted, so error rate is 2.76%.
* But 254/333, or 76% of defaulters missed.

## Classification Error Rates

* Terminology is from diagnostic testing in medicine.
    + E.G., pap smear to screen for cervical cancer
* The test can be positive ($T^+$) or negative ($T^-$), and an individual
may have the disease ($D^+$) or not ($D^-$):
\begin{center}
\begin{tabular}{ll|c|c|}
& \multicolumn{1}{c}{}& \multicolumn{2}{c}{Disease Status} \\
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{$D^-$} & \multicolumn{1}{c}{$D^+$} \\ \cline{3-4}
Test & $T^-$ & True & False \\
Result & & Negative (TN) & Negative (FN)  \\ \cline{3-4}
& $T^+$ &  False & True  \\
& & Positive (FP) & Positive (TP) \\ \cline{3-4}
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{TN+FP} & \multicolumn{1}{c}{FN+TP} \\
\end{tabular}
\end{center}

* Replace test with prediction and true disease
classification with true classification.

## Sensitivity, Specificity, True- and False-Positive Rates

* The observed true positive rate (TPR)
is the proportion of $D^+$ who are TP, or TP/(TP+FN). 
    + TP/(TP+FN) is an estimate of $P(T^+ \mid D^+)$, known as the _sensitivity_ 
of the test. 
    + Sensitivity is only 79/333 or 23.7% for `Default` data.
* The observed true negative rate is the proportion of $D^-$ who are
TN, or TN/(TN+FP).
    + TN/(TN+FP) is an estimate of the true negative rate $P(T^- \mid D^-)$,
known as the _specificity_ of the test. Specificity
is 9645/9667 or 99.8% for `Default` data.
    + The complement, FP/(TN+FP), is an estimate of the false positive rate (FPR)
$P(T^+ \mid D^-)$. FPR is 22/9667 or 0.2% for `Default` data.


## Receiver Operating Characteristic (ROC) Curves

* The classification of a subject with
$X=x_0$ as $T^+$ or $T^-$  
is made by comparing the posterior probability
$Pr(T^+|X=x_0)$ to a threshold $t=1/2$;
e.g., $>t$ classify as $T^+$, $\leq t$ classify as $T^-$.
* Lowering the threshold will increase the number of $T^+$,
and hence the TPR, $P(T^+\mid D^+)$, and the 
FPR, $P(T^+\mid D^-)$.
* The ROC curve is a plot plot of 
the TPR _versus_ FPR as we vary $t$.
    + Try to get a sense of the trade-off between true- and false-positive rates, and possibly choose an "optimal" $t$.

## ROC for LDA Classifier on Default Example

\scriptsize

```{r}
library(AUC) # for the roc() function; install.packages("AUC") to install
library(broom) # for the tidy() function
posteriorYes <- preds$posterior[,"Yes"]
trueYes <- # require a binary factor with 1=Yes
  (Default$default=="Yes") %>% as.numeric() %>% factor() 
ROCres <- roc(posteriorYes,trueYes)
tidyROCres <- tidy(ROCres)
```

## ROC Curve

\scriptsize

```{r}
ggplot(tidyROCres,aes(x=fpr,y=tpr)) + geom_point(pch=".")
```

## ROC Curve for Logistic Regression 

\scriptsize

```{r}
lfit <- glm(default ~ .,data=Default,family=binomial())
posteriorYesLogistic <- fitted(lfit)
ROClogist <- tidy(roc(posteriorYesLogistic,trueYes))
ggplot(ROClogist,aes(x=fpr,y=tpr)) + geom_point(pch=".")
```

## ROC Interpretation

* The points on the plot represent the FPR/TPR combinations
for each threshold.
* The ideal test threshold would yield a TPR of one and a FPR of zero, and so would appear in the upper-left corner of the ROC plot.
    + We usually select the threshold that is closest to 
the upper-left corner.
    + No obvious "best" threshold in this example. Perhaps
    the threshold that gives TPR of about 0.875
    and FPR of about 0.125.
    
\scriptsize

```{r}
tidyROCres %>% 
  filter(fpr >= 0.1249,fpr <= 0.1251)
```

## Re-classify

\scriptsize

```{r}
n <- nrow(Default); thresh <- 0.0457
dclass <- rep("No",n); dclass[posteriorYes>thresh] <- "Yes"
Default <- data.frame(Default,prednew = dclass)
xtabs(~ prednew + default, data=Default)
```

\normalsize

* Sensitivity is much better (about 87.7%).
* Many more false-positives (1209) 

## Area Under Curve (AUC)

* The area under the ROC curve is a measure of 
the overall performance of the classifier. 
* If TPR jumps to near one with FPR remaining 
low, the performance is good. 
    + This would give AUC near 1.
* A poor classifier would do no better than the
null classifier, which for threshold $t$ randomly
assigns $t\times 100$% to be the success category
and has AUC=1/2.


\scriptsize

```{r}
auc(ROCres)
```

## Quadratic Discriminant Analysis (QDA)

* Back to the $p=1$ case.
* If we let the variances differ by class $k$,
then it can be shown that the Bayes classifier 
assigns an observation with $X=x$ to the class with 
the largest 
$$\delta_k(x) = -\frac{1}{2}\frac{(x-\mu_k)^2}{\sigma^2} - \frac{1}{2} \log \sigma_k^2 + \log \pi_k,$$
which is a quadtratic function of $x$.
* See the text, page 149, for the 
multivariate version.

## QDA for the Default Data

\scriptsize

```{r}
qq <- qda(default ~ student + income + balance, data=Default)
preds <- predict(qq)
Default <- data.frame(Default,predicted = preds$class)
xtabs( ~ predicted + default, data=Default)
```

* Same as LDA for these data.

## QDA _vs_ LDA: Bias _vs_ Variance

* When should we assume the $K$ classes 
have a common variance matrix?
* LDA has fewer parameters and so lower variance.
    + When $p$ is large relative to $n$, the number of 
    parameters in QDA can become large, and variance of
    the posterior probabilities large.
* On the other hand, if the variances really **are**
different, the LDA posterior probabilities can be biased.
* Recommendation from the text:
If the training data set is large, use 
QDA. If not, use LDA.

## QDA _vs_ LDA: Decision Boundaries

\begin{center}
\includegraphics[height=2in]{Fig4_9.pdf}
\end{center}

* Text, Figure 4.9: $p=2$, Bayes (purple dashed), 
LDA (black dotted), QDA (green solid).


## Comparison of Classification Methods

* Performance depends on the nature of the true
decision boundary (curve that separates
the two classes).
* One can show that logistic regression 
leads to linear decision boundaries, just like LDA
    + Difference is in how the methods estimate
    the parameters of the boundaries.
    + Oftern give very similar results, as in the
    `Default` data.
* QDA gives quadratic boundaries.
* The KNN classifier is non-parametric and 
so there are no restrictions on the decision boundary.
* Performance depends on the nature of the true
decision boundary.

