---
title: "Week 5 exercises"
author: "Brad McNeney"
date: '2018-10-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stock Market Data

We will cover the material in the Chapter 4 lab
on LDA, QDA and KNN.
The data set is called `Smarket` and is in the 
`ISLR` package. The data are for 1250 days in
2001-2005. The variables are the percentage returns
from the five previous trading days 
(`Lag1` to `Lag5`), the number of shares 
traded in billions on the previous day(`Volume`), 
and the percentage
return (`Today`) and whether the market
was up or down (`Direction`) on that day.
The idea is to try and predict the market
using returns and volumne from previous days.



```{r}
library(ISLR)
data(Smarket)
head(Smarket)
```

Unlike the text I will not use `attach()` to 
attach the `Smarket` data frame to the 
search list (so that variables in `Smarket` are 
automatically accessible). The reason is that
`attach()`ing variables can lead to confusion
about where the variables come from, and can
lead to conflicting versions of variables.

### Data summaries

There are no notable correlations between the 
percent return variables. If anything there
is a slight negative correlation between
returns on adjacent days.

```{r}
library(tidyverse)
SmReturns <- select(Smarket, Today, Lag1:Lag5)
round(cor(SmReturns),3)
```


We can see a trend towards higher trading 
volume over 2001 and 2002, a leveling off in
2003 and 2004, and a sharper increase in volume
in 2005. 

```{r}
Smarket <- mutate(Smarket, Day = 1:nrow(Smarket))
ggplot(Smarket,aes(x=Day,y=Volume,color=Year)) + geom_line() + 
  geom_smooth()
```

### Logistic Regression

We are to do a logistic regression with 
`Direction` as the binary response. We are told
to use `contrasts()` to find out how `Direction`
will be coded as a dummy variable. Alternatively
we could create the dummy variable ourselves.

```{r}
Smarket <- mutate(Smarket,
                  dDirection = as.numeric(Direction=="Up"))
head(select(Smarket,Direction,dDirection))
```

We fit a model using the previous returns and volume
to predict direction. None of the coefficients
looks significant. We could also assess the 
overall fit of our model with a likelihood
ratio test. This is the analog to the 
F-test for multiple regression. We will not 
discuss the details and you are not responsible
for likelihood ratio tests.

```{r}
sfit <- glm(Direction ~ Lag1+Lag2+Lag3+Lag4+Lag5+Volume,
            data=Smarket,
            family=binomial())
round(summary(sfit)$coefficients,3)
sfitNull <- glm(Direction ~ 1,
            data=Smarket,family=binomial())
anova(sfitNull,sfit)
pchisq(3.5907,df=6,lower.tail=FALSE)
```


Though the model terms
are not significant, we might still try
to interpret them.
For example, the negative coefficient for `Lag1` tells
us that an increase in the Lag1 returns 
corresponds to a decrease in the log-odds of the market going up the next day, and therefore a
decrease in the probability of the market going up.
That is, a strong day
on the market the day before makes a positive return
less likely today. 
This is consistent 
with our observation that the correlation 
between returns on adjacent days is negatively correlated.

We can use `predict()` to make predictions using
data on previous days and assess 
the prediction error.

```{r}
predDir <- function(fit,dat) {
  probs <- predict(fit,newdata=dat,type="response")
  n <- nrow(dat) # Data used to fit
  pp <- rep("Down",n)
  pp[probs>0.5] <- "Up"
  pp
}
Smarket <- mutate(Smarket,predDirection = predDir(sfit,Smarket))
xtabs(~ predDirection+Direction, data=Smarket)
with(Smarket,mean(predDirection != Direction))
```

The  prediction error of 48% is optimistic because
we are making predictions on the same data that
was used to fit the model. To get a test set error
we need to hold out some of the data. The text
uses the 2005 data as a hold out. 

```{r}
train <- (Smarket$Year < 2005)
Smarket.train <- Smarket[train,]
Smarket.test <- Smarket[!train,]
sfit <- glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,
            data=Smarket.train, family=binomial())
Smarket.test <- mutate(Smarket.test, 
                       predDirection = predDir(sfit,Smarket.test))
xtabs(~predDirection + Direction,data=Smarket.test)
with(Smarket.test,mean(predDirection != Direction)) 
```

We see a higher error rate than 50%, which is
the rate we'd expect from just guessing whether
the market will go up or down.

None of the model terms look very significant
but especially those for volume and lags of 3 or more days. 
Remove these and re-train, etc.

```{r}
round(summary(sfit)$coefficients,3)
sfit <- glm(Direction~Lag1+Lag2,
            data=Smarket.train, family=binomial())
Smarket.test <- mutate(Smarket.test, 
                       predDirection = predDir(sfit,Smarket.test))
xtabs(~predDirection + Direction,data=Smarket.test)
with(Smarket.test,mean(predDirection != Direction)) 
```

The reduced model appears to be a little better
than guessing.

### Linear Discriminant Analysis

```{r}
library(MASS)
sfit <- lda(Direction ~ Lag1+Lag2, data=Smarket.train)
preds <- predict(sfit,newdata=Smarket.test)$class
Smarket.test <- mutate(Smarket.test,predDirection = preds)
xtabs(~predDirection + Direction,data=Smarket.test)
with(Smarket.test,mean(predDirection != Direction)) 
```

Same predictions and error rate as logistic regression.


### Quadratic Discriminant Analysis

```{r}
sfit <- qda(Direction ~ Lag1+Lag2, data=Smarket.train)
preds <- predict(sfit,newdata=Smarket.test)$class
Smarket.test <- mutate(Smarket.test,predDirection = preds)
xtabs(~predDirection + Direction,data=Smarket.test)
with(Smarket.test,mean(predDirection != Direction)) 
```

The error rate is about 40%, the best so far.

### K-Nearest Neighbors

```{r}
library(class) # help(knn)
# A function called select() is in the MASS package,
# and will be used before select() from dplyr.
# Use :: to specifically use select from dplyr
train.X <- dplyr::select(Smarket.train,Lag1,Lag2)
test.X <- dplyr::select(Smarket.test,Lag1,Lag2)
train.Direction <- Smarket.train$Direction
set.seed(1) # for randomly breaking neighbor ties
preds <- knn(train=train.X,test=test.X,cl=train.Direction,k=3)
Smarket.test <- mutate(Smarket.test,predDirection = preds)
xtabs(~predDirection + Direction,data=Smarket.test)
with(Smarket.test,mean(predDirection != Direction)) 
```

The prediction error of KNN is the highest of 
the methods considered.

