---
title: "Week 13 Exercises"
author: "Brad McNeney"
date: '2018-11-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will cluster on the US Arrests dataset that
comes with R.
There is a short quiz on Canvas named
`Week 13 Tutorial Quiz` for you to fill out after doing these questions.


```{r}
data(USArrests)
head(USArrests)
```

1. **Without any scaling or centring of the data** perform hierarchical clustering with 
complete linkage and Euclidean distance.
Then cut the tree at the height that 
produces three clusters and report the 
states in each cluster.


2. Now **centre and scale** the variables to 
have mean 0 and SD 1 and perform the clustering
with complete linkage and Euclidean distance.
Cut the tree at the height that 
produces three clusters and report the 
states in each cluster.


3. Using the centred and scaled data, compute the
first two PCs. Do a scatterplot of the first
two PC scores for each of the US states and
colour the points by the three-cluster assignments
from question 2.

4. Using the centred and scaled data, 
perform the clustering
with single linkage and Manhattan distance.
Cut the tree at the height that 
produces three clusters and report the 
states in each cluster.


5.  Do a scatterplot of the first
two PC scores for each of the US states and
colour the points by the three-cluster assignments
from question 4.
